---
title: "glmm.pcn.reduced"
output: html_document
date: "2025-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(glmmTMB)
library(tidyverse)
library(ICC)
library(emmeans)
library(dendextend)
library(performance)
```

```{r}
set2.5backup <- read_delim("set2.5_20250324_safety.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

make all the SC the same
```{r}
set2.5backup$centername <- set2.5backup$`Center Name` %>%
  str_replace("WELLCOME SANGER INSTITUTE", "SC") %>%
  str_replace("THE WELLCOME TRUST SANGER INSTITUTE", "SC")
```

```{r}
setMC <- set2.5backup %>%
  filter(PCN > 2)

setpostMC <- set2.5backup %>%
  filter(PCN > 2) %>%
  filter(prepost == "post")
```

```{r}
setpostMC %>%
  select(BioProject, centername, Instrument) %>%
  group_by(BioProject, centername, Instrument) %>%
  mutate(count3 = n()) %>%
  group_by(BioProject) %>%
  mutate(countBP = n()) %>%
  group_by(BioProject, Instrument) %>%
  mutate(BPins = n()) %>%
  group_by(BioProject, centername) %>%
  mutate(BPcen = n()) %>%
  filter(countBP != count3) %>%
  ungroup() %>%
  distinct(BioProject)
```
2 BioProjects were sequenced on more than 2 models of instrument, 1 may have been (just wasn't specific enough)
Each BioProject was sequenced at 1 seqcenter max -- no bioprojects spread across sequencing centers
if we account for bioproject, don't worry about sequencing center

should include instrument as a random effect - do not expect strain to have an effect on this. time might, as that might influence what reagents/prep is used

within each bioproject (random), do we see an effect of strain (fixed)
or rather, I should say,
within a strain (fixed), do we see an effect of bioproject (random)

```{r}
m9 = glmmTMB(PCN ~ as.numeric(year) + strain + 
          (1 | strain:BioProject),
        data = setpostMC,
             family = Gamma(link="log"))
summary(m9)
```

How many BioProjects does each strain have...?

```{r}
setpostMC %>%
  select(strain, BioProject) %>%
  distinct() %>%
  group_by(strain) %>%
  mutate(bpct = n()) %>%
  ungroup() %>%
  select(strain, bpct) %>%
  distinct() %>%
  arrange(bpct)
```
Some strains only have 1 BioProject, some have more

```{r}
setpostMCgt10 <- setpostMC %>%
  group_by(strain) %>%
  mutate(strainCount = n()) %>%
  select(strain, strainCount) %>%
  distinct() %>%
  #arrange(desc(strainCount))
  filter(strainCount > 10)
```

```{r}
set395 <- left_join(setpostMCgt10, setpostMC, by = "strain")
```


```{r}
m10 = glmmTMB(PCN ~ as.numeric(year) + strain + 
          (1 | strain:BioProject),
        data = set395,
             family = Gamma(link="log"))
summary(m10)
```
Okay so this might be a tough sell.......

Let's plot it and look
```{r}
set395 %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free") +
  theme(legend.position="none") 
```

let's just look at effect of multicopy PCN overtime by strain
```{r}
m11 = glmmTMB(PCN ~ as.numeric(year) + strain + (1 | strain:BioProject),
        data = setMC,
             family = Gamma(link="log"))
summary(m11)
```

```{r}
setMC$numYear <- setMC$year %>%
  as.numeric()
```


```{r}
m17 = glmmTMB(PCN ~ numYear + strain + (1 | numYear:BioProject) +
                (1 | strain:BioProject),
        data = setMC,
             family = Gamma(link="log"))
summary(m17)
```

```{r}
m12 = glmmTMB(PCN ~ as.numeric(year) + strain,
        data = setMC,
             family = Gamma(link="log"))

m13 = glmmTMB(PCN ~ as.numeric(year) + strain +
                (1|Instrument) + (1|centername),
        data = setMC,
             family = Gamma(link="log"))

m14 = glmmTMB(PCN ~ as.numeric(year) + strain + (1|centername),
        data = setMC,
             family = Gamma(link="log"))
```


```{r}
#anova(m11, m13, m12)
#anova(m11, m12) # m11 is better than simplest (m12)
anova(m11, m14, m12) 
```

```{r changing model}
m15 = glmmTMB(PCN ~ as.numeric(year) + strain,
        data = setMC,
             family = lognormal(link = "log"))

m16 = glmmTMB(PCN ~ as.numeric(year) + strain +
                (1|strain:BioProject),
        data = setMC,
             family = lognormal(link = "log"))

summary(m15)
summary(m16)
```

# learning about bioproject size
```{r}
bbp <- setMC %>%
  select(BioProject#, strain
         ) %>%
  group_by(BioProject) %>%
  mutate(count = n()) %>%
  distinct() %>%
  arrange(desc(count)) %>%
  filter(count >= 10)
```

```{r}
left_join(bbp, setMC, by = "BioProject") %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point()
```

```{r early big bioprojects}
left_join(bbp, setMC, by = "BioProject") %>%
  filter(prepost == "pre") %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point()
```

```{r}
bbbp <- left_join(bbp, setMC, by = "BioProject")
```

```{r}
m18 = glmmTMB(PCN ~ numYear + strain +
                (1|strain:BioProject) +
                (numYear|BioProject),
        data = bbbp,
             family = Gamma(link="log"))
summary(m18)
```

```{r}
m19 = glmmTMB(PCN ~ numYear + strain +
                (1|strain:BioProject) +
                (1|numYear:BioProject),
        data = bbbp,
             family = Gamma(link="log"))
summary(m19)
```

should I assess evenness of BioProjects...?

# trying again with large strains within large bioprojects

```{r}
topstrainlist <- bbbp %>%
  ungroup() %>%
  select(strain) %>%
  group_by(strain) %>%
  mutate(nstrain = n()) %>%
  distinct()%>%
  arrange(desc(nstrain)) %>%
  filter(nstrain >=10)
```

```{r}
topstrainbp <- left_join(topstrainlist, bbbp, by = "strain")

topstrainbp %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point()

topstrainbp %>%
  ungroup() %>%
  group_by(BioProject) %>%
  mutate(bpcount = n()) %>%
  filter(bpcount >= 10) %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free")

topstrainbp %>%
  ungroup() %>%
  group_by(BioProject) %>%
  mutate(bpcount = n()) %>%
  filter(bpcount >= 10) %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point() +
  facet_wrap(~BioProject, scales = "free")
```

```{r}
m20 = glmmTMB(PCN ~ numYear + strain +
                (1|strain:BioProject) +
                (1|numYear:BioProject),
        data = topstrainbp,
             family = Gamma(link="log"))
summary(m20)
```


let's say - only include strain/bioproject sets that have >5 samples in that strain/bioproject...?

```{r pcn by year/strain/bioproject for top strain/bp combos}
setMC %>%
  select(strain, BioProject) %>%
  group_by(strain, BioProject) %>%
  mutate(count = n()) %>%
  distinct() %>%
  filter(count >= 5) %>%
  left_join(., setMC, by = c("strain", "BioProject")) %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free")
```

```{r set with top strains and bp}
oe5 <- setMC %>%
  select(strain, BioProject) %>%
  group_by(strain, BioProject) %>%
  mutate(count = n()) %>%
  distinct() %>%
  filter(count >= 5) %>%
  ungroup() %>%
  left_join(., setMC, by = c("strain", "BioProject"))
```

```{r palette}
strain.c11 <- c(
  "S99.5_10" = "dodgerblue2", #1
  "S99.5_11" = "palegreen2",
  "S99.5_15" = "orchid1",
  "S99.5_17" = "blue1",
  "S99.5_2" = "#E31A1C", # red, 2
  "S99.5_20" = "darkturquoise",
  "S99.5_4" = "#CAB2D6", # lt purple
  "S99.5_49" = "#FB9A99",
  "S99.5_5" = "#FF7F00", # orange, 5
  "S99.5_7" = "gold1",
  "S99.5_8" = "skyblue2"
)
```


```{r pcn plots}
oe5 %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point() +
  scale_y_sqrt() +
  theme_bw() +
  # geom_smooth(method = "lm") + # too many slopes -> too messy
  xlab("Year of Collection") +
  # scale_color_viridis(discrete=TRUE, option="inferno")
  scale_color_manual(values = strain.c11)

oe5 %>%
  filter(PCN < 50) %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point()

oe5 %>%
  filter(prepost=="post") %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free")
```

```{r factor & relevel}
oe5$strainFac <- factor(oe5$strain)

oe5$strainFac <- relevel(oe5$strainFac, ref = "S99.5_8")
```

```{r oe5 by strain and bioproject boxplot}
oe5 %>%
  ggplot(aes(x = forcats::fct_reorder(strainFac, PCN, .fun = median),
             y = PCN, color = BioProject)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_sqrt() +
  labs(x = "Strain (colored by BioProject)") #+
  #scale_color_viridis(discrete=TRUE, option="inferno")
```

```{r}
oe5 %>%
  mutate(avgPCN = mean(PCN)) %>%
  group_by(strain) %>%
  mutate(strainMean = mean(PCN),
         straincount = n(),
         strainSD = sd(PCN)) %>%
  select(strain, strainMean, strainSD, avgPCN, straincount) %>%
  distinct() %>%
  arrange(desc(straincount))
```

```{r}
m21 = glmmTMB(PCN ~ numYear + strainFac +
                (1|strainFac:BioProject) +
                (1|numYear:BioProject), # AIC is lower with them both included 
        data = oe5,
             family = Gamma(link="log"))
summary(m21)
```

```{r}
emmeans(m21, ~strainFac)

pairs(emmeans(m21, ~strainFac), adjust = "tukey")

emmeans(m21, pairwise ~strainFac)
```

```{r median PCN in pre vs post mc isolates}
setMC %>%
  group_by(prepost) %>%
  mutate(medPCN = median(PCN)) %>%
  select(prepost, medPCN) %>%
  distinct()
```

```{r}
performance(m21)
r2(m21)
icc(m21)
```


```{r}
emm <- emmeans(m21, ~ strainFac)

emm

plot(emm, comparisons = TRUE, adjust = "tukey")
contrast(emm, method = "eff")  # compares each level to the average

emmeans(m21, pairwise ~ strainFac, adjust = "bh")
contrast(emm, method = "trt.vs.ctrl", adjust = "bh")
```

| Strain    | Log(PCN) `emmean` | Back-transformed PCN (`exp(emmean)`) |
| --------- | ----------------- | ------------------------------------ |
| S99.5\_7  | 3.45              | \~31.5                               |
| S99.5\_15 | 3.25              | \~25.8                               |
| S99.5\_2  | 3.29              | \~26.8                               |
| S99.5\_49 | 1.81              | \~6.1                                |
| ...       | ...               | ...                                  |

```{r}
summary(emm, type = "response")
```

```{r how many strains in the n=455 df}
oe5 %>%
  select(strain) %>%
  distinct()
```
