---
title: "glmm.pcn.reduced"
output: html_document
date: "2025-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(glmmTMB)
library(tidyverse)
library(ICC)
library(emmeans)
library(dendextend)
library(performance)
```

```{r}
set2.5backup <- read_delim("set2.5_20250324_safety.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

make all the SC the same
```{r}
set2.5backup$centername <- set2.5backup$`Center Name` %>%
  str_replace("WELLCOME SANGER INSTITUTE", "SC") %>%
  str_replace("THE WELLCOME TRUST SANGER INSTITUTE", "SC")
```

```{r}
setMC <- set2.5backup %>%
  filter(PCN > 2)

setpostMC <- set2.5backup %>%
  filter(PCN > 2) %>%
  filter(prepost == "post")
```

```{r}
setpostMC %>%
  select(BioProject, centername, Instrument) %>%
  group_by(BioProject, centername, Instrument) %>%
  mutate(count3 = n()) %>%
  group_by(BioProject) %>%
  mutate(countBP = n()) %>%
  group_by(BioProject, Instrument) %>%
  mutate(BPins = n()) %>%
  group_by(BioProject, centername) %>%
  mutate(BPcen = n()) %>%
  filter(countBP != count3) %>%
  ungroup() %>%
  distinct(BioProject)
```
2 BioProjects were sequenced on more than 2 models of instrument, 1 may have been (just wasn't specific enough)
Each BioProject was sequenced at 1 seqcenter max -- no bioprojects spread across sequencing centers
if we account for bioproject, don't worry about sequencing center

should include instrument as a random effect - do not expect strain to have an effect on this. time might, as that might influence what reagents/prep is used

within each bioproject (random), do we see an effect of strain (fixed)
or rather, I should say,
within a strain (fixed), do we see an effect of bioproject (random)

```{r}
m9 = glmmTMB(PCN ~ as.numeric(year) + strain + 
          (1 | strain:BioProject),
        data = setpostMC,
             family = Gamma(link="log"))
summary(m9)
```

How many BioProjects does each strain have...?

```{r}
setpostMC %>%
  select(strain, BioProject) %>%
  distinct() %>%
  group_by(strain) %>%
  mutate(bpct = n()) %>%
  ungroup() %>%
  select(strain, bpct) %>%
  distinct() %>%
  arrange(bpct)
```
Some strains only have 1 BioProject, some have more

```{r}
setpostMCgt10 <- setpostMC %>%
  group_by(strain) %>%
  mutate(strainCount = n()) %>%
  select(strain, strainCount) %>%
  distinct() %>%
  #arrange(desc(strainCount))
  filter(strainCount > 10)
```

```{r}
set395 <- left_join(setpostMCgt10, setpostMC, by = "strain")
```


```{r}
m10 = glmmTMB(PCN ~ as.numeric(year) + strain + 
          (1 | strain:BioProject),
        data = set395,
             family = Gamma(link="log"))
summary(m10)
```
Okay so this might be a tough sell.......

Let's plot it and look
```{r}
set395 %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free") +
  theme(legend.position="none") 
```

let's just look at effect of multicopy PCN overtime by strain
```{r}
m11 = glmmTMB(PCN ~ as.numeric(year) + strain + (1 | strain:BioProject),
        data = setMC,
             family = Gamma(link="log"))
summary(m11)
```

```{r}
setMC$numYear <- setMC$year %>%
  as.numeric()
```


```{r}
m17 = glmmTMB(PCN ~ numYear + strain + (1 | numYear:BioProject) +
                (1 | strain:BioProject),
        data = setMC,
             family = Gamma(link="log"))
summary(m17)
```

```{r}
m12 = glmmTMB(PCN ~ as.numeric(year) + strain,
        data = setMC,
             family = Gamma(link="log"))

m13 = glmmTMB(PCN ~ as.numeric(year) + strain +
                (1|Instrument) + (1|centername),
        data = setMC,
             family = Gamma(link="log"))

m14 = glmmTMB(PCN ~ as.numeric(year) + strain + (1|centername),
        data = setMC,
             family = Gamma(link="log"))
```


```{r}
#anova(m11, m13, m12)
#anova(m11, m12) # m11 is better than simplest (m12)
anova(m11, m14, m12) 
```

```{r changing model}
m15 = glmmTMB(PCN ~ as.numeric(year) + strain,
        data = setMC,
             family = lognormal(link = "log"))

m16 = glmmTMB(PCN ~ as.numeric(year) + strain +
                (1|strain:BioProject),
        data = setMC,
             family = lognormal(link = "log"))

summary(m15)
summary(m16)
```

# learning about bioproject size
```{r}
bbp <- setMC %>%
  select(BioProject#, strain
         ) %>%
  group_by(BioProject) %>%
  mutate(count = n()) %>%
  distinct() %>%
  arrange(desc(count)) %>%
  filter(count >= 10)
```

```{r}
left_join(bbp, setMC, by = "BioProject") %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point()
```

```{r early big bioprojects}
left_join(bbp, setMC, by = "BioProject") %>%
  filter(prepost == "pre") %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point()
```

```{r}
bbbp <- left_join(bbp, setMC, by = "BioProject")
```

```{r}
m18 = glmmTMB(PCN ~ numYear + strain +
                (1|strain:BioProject) +
                (numYear|BioProject),
        data = bbbp,
             family = Gamma(link="log"))
summary(m18)
```

```{r}
m19 = glmmTMB(PCN ~ numYear + strain +
                (1|strain:BioProject) +
                (1|numYear:BioProject),
        data = bbbp,
             family = Gamma(link="log"))
summary(m19)
```

should I assess evenness of BioProjects...?

# trying again with large strains within large bioprojects

```{r}
topstrainlist <- bbbp %>%
  ungroup() %>%
  select(strain) %>%
  group_by(strain) %>%
  mutate(nstrain = n()) %>%
  distinct()%>%
  arrange(desc(nstrain)) %>%
  filter(nstrain >=10)
```

```{r}
topstrainbp <- left_join(topstrainlist, bbbp, by = "strain")

topstrainbp %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point()

topstrainbp %>%
  ungroup() %>%
  group_by(BioProject) %>%
  mutate(bpcount = n()) %>%
  filter(bpcount >= 10) %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free")

topstrainbp %>%
  ungroup() %>%
  group_by(BioProject) %>%
  mutate(bpcount = n()) %>%
  filter(bpcount >= 10) %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point() +
  facet_wrap(~BioProject, scales = "free")
```

```{r}
m20 = glmmTMB(PCN ~ numYear + strain +
                (1|strain:BioProject) +
                (1|numYear:BioProject),
        data = topstrainbp,
             family = Gamma(link="log"))
summary(m20)
```


let's say - only include strain/bioproject sets that have >5 samples in that strain/bioproject...?

```{r pcn by year/strain/bioproject for top strain/bp combos}
setMC %>%
  select(strain, BioProject) %>%
  group_by(strain, BioProject) %>%
  mutate(count = n()) %>%
  distinct() %>%
  filter(count >= 5) %>%
  left_join(., setMC, by = c("strain", "BioProject")) %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free")
```

```{r set with top strains and bp}
oe5 <- setMC %>%
  select(strain, BioProject) %>%
  group_by(strain, BioProject) %>%
  mutate(count = n()) %>%
  distinct() %>%
  filter(count >= 5) %>%
  ungroup() %>%
  left_join(., setMC, by = c("strain", "BioProject"))
```

```{r palette}
strain.c11 <- c(
  "S99.5_10" = "dodgerblue2", #1
  "S99.5_11" = "palegreen2",
  "S99.5_15" = "orchid1",
  "S99.5_17" = "blue1",
  "S99.5_2" = "#E31A1C", # red, 2
  "S99.5_20" = "darkturquoise",
  "S99.5_4" = "#CAB2D6", # lt purple
  "S99.5_49" = "#FB9A99",
  "S99.5_5" = "#FF7F00", # orange, 5
  "S99.5_7" = "gold1",
  "S99.5_8" = "skyblue2"
)
```


```{r pcn plots}
oe5 %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point() +
  scale_y_sqrt() +
  theme_bw() +
  # geom_smooth(method = "lm") + # too many slopes -> too messy
  xlab("Year of Collection") +
  # scale_color_viridis(discrete=TRUE, option="inferno")
  scale_color_manual(values = strain.c11)

oe5 %>%
  filter(PCN < 50) %>%
  ggplot(aes(x = numYear, y = PCN, color = strain)) +
  geom_point()

oe5 %>%
  filter(prepost=="post") %>%
  ggplot(aes(x = numYear, y = PCN, color = BioProject)) +
  geom_point() +
  facet_wrap(~strain, scales = "free")
```

```{r factor & relevel}
oe5$strainFac <- factor(oe5$strain)

oe5$strainFac <- relevel(oe5$strainFac, ref = "S99.5_8")
```

```{r oe5 by strain and bioproject boxplot}
oe5 %>%
  ggplot(aes(x = forcats::fct_reorder(strainFac, PCN, .fun = median),
             y = PCN, color = BioProject)) +
  geom_boxplot() +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_sqrt() +
  labs(x = "Strain (colored by BioProject)") #+
  #scale_color_viridis(discrete=TRUE, option="inferno")
```

```{r}
oe5 %>%
  mutate(avgPCN = mean(PCN)) %>%
  group_by(strain) %>%
  mutate(strainMean = mean(PCN),
         straincount = n(),
         strainSD = sd(PCN)) %>%
  select(strain, strainMean, strainSD, avgPCN, straincount) %>%
  distinct() %>%
  arrange(desc(straincount))
```

```{r winning glmm}
m21 = glmmTMB(PCN ~ numYear + strainFac +
                (1|strainFac:BioProject) +
                (1|numYear:BioProject), # AIC is lower with them both included 
        data = oe5,
             family = Gamma(link="log"))
summary(m21)
```

```{r}
emmeans(m21, ~strainFac)

pairs(emmeans(m21, ~strainFac), adjust = "tukey")

emmeans(m21, pairwise ~strainFac)
```

```{r median PCN in pre vs post mc isolates}
setMC %>%
  group_by(prepost) %>%
  mutate(medPCN = median(PCN)) %>%
  select(prepost, medPCN) %>%
  distinct()
```

```{r}
performance(m21)
r2(m21)
icc(m21)
```


```{r}
emm <- emmeans(m21, ~ strainFac)

emm

plot(emm, comparisons = TRUE, adjust = "tukey")
contrast(emm, method = "eff")  # compares each level to the average

emmeans(m21, pairwise ~ strainFac, adjust = "bh")
contrast(emm, method = "trt.vs.ctrl", adjust = "bh")
```

| Strain    | Log(PCN) `emmean` | Back-transformed PCN (`exp(emmean)`) |
| --------- | ----------------- | ------------------------------------ |
| S99.5\_7  | 3.45              | \~31.5                               |
| S99.5\_15 | 3.25              | \~25.8                               |
| S99.5\_2  | 3.29              | \~26.8                               |
| S99.5\_49 | 1.81              | \~6.1                                |
| ...       | ...               | ...                                  |

```{r}
summary(emm, type = "response")
```

```{r how many strains in the n=455 df}
oe5 %>%
  select(strain) %>%
  distinct()
```

##checking instrument in oe5

Are all bioprojects in this set sequenced on 1 machine?
```{r}
oe5 %>%
  select(BioProject, Instrument) %>%
  distinct() %>%
  group_by(BioProject) %>%
  mutate(uniqInstrument = n())
```
Some bioprojects were sequenced on 2 machines, most were sequenced on 1

## oe5 how big are the bioprojects?

```{r}
oe5 %>%
  group_by(BioProject) %>%
  mutate(size = n()) %>%
  select(BioProject, size) %>%
  distinct() %>%
  arrange(desc(size))
```
2 bioprojects account for >50% of the samples in the oe5 dataset

what do their time & PCN trends look like?

```{r}
oe5 %>%
  filter(BioProject == "PRJEB2812" | BioProject == "PRJEB9575") %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             shape = BioProject, color = strain)) +
  geom_point()
```
okay, so one of the big bioprojects is all the early samples
this could be good but also bad?
when were these sequenced...?

```{r}
oe5 %>%
  filter(BioProject == "PRJEB2812") %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = strain)) +
  geom_point()

## we don't have date of sequencing in this dataset
## would need to go back to the other one...
```

we may need to look at some of the papers themselves...
from SRA
matching projects with publications
PRJEB2812 - https://pubmed.ncbi.nlm.nih.gov/24713324/
PRJEB9575 - no paper
PRJNA418899 - no paper
PRJNA324190 - no paper
PRJEB2686 - https://pubmed.ncbi.nlm.nih.gov/26873713/

```{r}
setMC %>%
  filter(as.numeric(year) < 1995) %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = BioProject)) +
  geom_point()

## also check: do these bioprojects have some lower copy isolates (later)?
# evidence for this to be true vs untrue?
# we can't find obvious sign of technical issue (will look in revisions)
# multiple bioprojects have similar ranges of pcn (mention this to reviewer)
# there is evidence that this is not a technical effect, but we will investigate in revisions.
```
SEVERAL BioProjects in similar range!!

BioProjects...
```{r}
set2.5backup %>%
  filter(BioProject == "PRJEB131165" |
           BioProject == "PRJEB14816" |
           BioProject == "PRJEB2812" |
           BioProject == "PRJEB2998" |
           BioProject == "PRJNA228475" |
           BioProject == "PRJNA231221" |
           BioProject == "PRJNA267633" |
           BioProject == "PRJNA374337") %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = BioProject)) +
  geom_point() +
  theme_bw() +
  xlab("Year of Collection")
```

Multiple BioProjects find the same result, where earlier samples have higher copy number compared to later samples. We expect consistency of methods within a BioProject

# getting info on the bioprojects
I'm coming back now to get the BioProjects for multicopy plasmids
I need to get info on the extraction technique
(also get date of sequencing -- but that's for later)

```{r}
setMC %>%
  select(BioProject) %>%
  distinct() # 75 BioProjects have a multicopy plasmid

setMC %>%
  select(BioProject, prepost) %>%
  distinct() %>%
  group_by(BioProject) %>%
  mutate(count = n()) %>%
  filter(count > 1) %>% # 3 bioprojects have samples both pre/post
  distinct(BioProject)
```
The 3 bioprojects pre/post are
PRJEB14816
PRJEB13165
PRJNA374337

```{r}
setMC %>%
  filter(BioProject == "PRJEB14816" | BioProject == "PRJEB13165" |
           BioProject == "PRJNA374337") %>%
  ggplot(aes(x = as.numeric(year), color = BioProject, y = PCN)) +
  geom_point()
```
Withink the 3 BioProjects that cross the 1995 line, they all show substantial decline in multicopy PCN over time

This isn't enough points to really do statistics but is certainly a start

I think we do need to look at the papers
It is the best use of time to look at the largest bioprojects 1st

```{r}
setMC %>%
  group_by(BioProject) %>%
  mutate(BPsize = n()) %>%
  select(BioProject, BPsize) %>%
  distinct() %>%
  arrange(desc(BPsize))
```
matching projects with publications - top 5
PRJEB2812 - https://pubmed.ncbi.nlm.nih.gov/24713324/ - dordel et al 2014, extraction technique not clear
PRJEB9575 - found with web search, https://pmc.ncbi.nlm.nih.gov/articles/PMC5573676/ - Moradigaravand et al 2017, QIAxtractor (Qiagen)
PRJNA418899 - found with web search, https://elifesciences.org/articles/42486 - Young et al 2019, mechanical lysis step (FastPrep; MPBiomedicals) followed by a commercial kit (QuickGene; Autogen Inc)
PRJNA324190 - web search finds a paper that uses the data https://pmc.ncbi.nlm.nih.gov/articles/PMC11730646/
PRJEB2686 - https://pubmed.ncbi.nlm.nih.gov/26873713/ -- chang et al 2016, QIAamp DNA Mini Kit
(Qiagen)

Next 5 bioprojects
PRJEB2998	- not findable			
PRJEB2655 - 5 papers (https://www.ncbi.nlm.nih.gov/bioproject/?term=(PRJEB2655)+AND+bioproject_sra%5Bfilter%5D+NOT+bioproject_gap%5Bfilter%5D) - 1 "prev described", 4 Genomic DNA was extracted using the MasterPure™ Gram Positive DNA Purification Kit (Cambio, Dry Drayton, UK)
PRJEB2870 - https://pmc.ncbi.nlm.nih.gov/articles/PMC4020051 - Uhleman et al	, QIAamp DNA Mini Kit (Qiagen)			
PRJNA368710 - found with web search, https://www.sciencedirect.com/science/article/pii/S2451945617300971 - wang et al 2017 - DNA Blood and Tissue Kit 69504 Qiagen	
PRJNA369475	- found with web search, https://elifesciences.org/articles/30637 - young et al 2017, DNA extracted using a mechanical lysis step (FastPrep; MPBiomedicals) followed by a commercial kit (QuickGene; Fujifilm)

Let's just look at all bioprojects with at least 10 mc pt181
PRJEB11627 - found with web search, https://pmc.ncbi.nlm.nih.gov/articles/PMC4809937/ - Strauß et al 2016, Staphylococcal chromosomal DNA was extracted using the MagAttract HMW DNA kit (Qiagen) according to the manufacturer's instructions for Gram-positive bacteria.
PRJEB11177 - found with web search, https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1001944 - Auguet et al 2016, methods from hughes et al 2015 (https://academic.oup.com/jac/article-abstract/70/12/3191/2363662?redirectedFrom=fulltext&login=true) - extraction technique unclear
PRJNA445516 - found with web search, https://pmc.ncbi.nlm.nih.gov/articles/PMC6113501/ - Mason et al 2018, extraction technique not clear	

also need to prioritize looking for papers from the projects with early seq
```{r}
setMC %>%
  filter(prepost == "pre") %>%
  group_by(BioProject) %>%
  mutate(BPsize = n()) %>%
  select(BioProject, BPsize) %>%
  distinct() %>%
  arrange(desc(BPsize))
```
PRJEB2812	- found with web search, HARKINS MRSA PAPER - Qiagen QIAcube
PRJEB2998	- not findable			
PRJNA374337 - found with abstract search, https://pmc.ncbi.nlm.nih.gov/articles/PMC5932376/ - extraction technique not clear
PRJEB14816 - found with web search, https://pmc.ncbi.nlm.nih.gov/articles/PMC5724248/ - extraction technique not clear
PRJNA231221 - fda-argos, https://www.ncbi.nlm.nih.gov/pubmed/31346170 - extraction technique not readily apparent
PRJNA267633 - https://www.ncbi.nlm.nih.gov/pubmed/26112789 - they cite their extraction technique from this paper which is about trees (https://www.afs-journal.org/articles/forest/abs/1999/03/AFS_0003-4312_1999_56_3_ART0008/AFS_0003-4312_1999_56_3_ART0008.html)
PRJNA228475 - not findable
PRJEB13165 - found with web search, https://pmc.ncbi.nlm.nih.gov/articles/PMC12691627/ & https://journals.asm.org/doi/10.1128/mbio.01755-19 - extraction technique unclear

let's look at the bioprojects we have extraction info for
```{r}
setMC %>%
  filter(BioProject == "PRJEB9575" |
           BioProject == "PRJNA418899" |
           BioProject == "PRJEB2686" |
           BioProject == "PRJEB2655" |
           BioProject == "PRJEB2870" |
           BioProject == "PRJNA368710" |
           BioProject == "PRJNA369475" |
           BioProject == "PRJEB11627" |
           BioProject == "PRJEB2812") %>%
  ggplot(aes(x = as.numeric(year),
             color = BioProject,
             y = PCN)) +
  geom_point()

setMC %>%
  filter(BioProject == "PRJEB9575" | # qiaxtractor
           BioProject == "PRJNA418899" | #fastprep/quickgene
           BioProject == "PRJEB2686" | # QIAamp DNA Mini Kit (Qiagen)
           BioProject == "PRJEB2655" | # masterpure
           BioProject == "PRJEB2870" | # QIAamp DNA Mini Kit (Qiagen)
           BioProject == "PRJNA368710" | # DNA Blood and Tissue Kit 69504 Qiagen
           BioProject == "PRJNA369475" | #fastprep/quickgene
           BioProject == "PRJEB11627" | # MagAttract HMW DNA kit (Qiagen) 
           BioProject == "PRJEB2812") %>% # Qiagen QIAcube
  filter(prepost == "pre") %>%
  ggplot(aes(x = as.numeric(year),
             color = BioProject,
             y = PCN)) +
  geom_point() ## there is only 1 bioproject with old samples for that we have extraction technique for
```

the old project uses Qiagen QIAcube
chemistry for QIAamp DNA Mini Kit & QIAcube are the same (allegedly)
let's look at projects using those kits

```{r}
setMC %>%
  filter(BioProject == "PRJEB2686" | # QIAamp DNA Mini Kit (Qiagen)
           BioProject == "PRJEB2870" | # QIAamp DNA Mini Kit (Qiagen)
           BioProject == "PRJNA368710" | # DNA Blood and Tissue Kit 69504 Qiagen
           BioProject == "PRJEB2812") %>% # Qiagen QIAcube
  ggplot(aes(x = as.numeric(year),
             color = BioProject,
             y = PCN)) +
  geom_point()
```

```{r the 3 bioproj that cross the line}
setMC %>%
  filter(BioProject == "PRJEB14816" | # extraction technique not clear
           BioProject == "PRJEB13165" | # extraction technique not clear
           BioProject == "PRJNA374337") %>% # extraction technique not clear
  ggplot(aes(x = as.numeric(year), color = strain, y = PCN,
             shape = centername)) +
  geom_point()
```

looking at just the recent bioproject
```{r}
setMC %>%
  filter(BioProject == "PRJEB9575" | # qiaxtractor
           BioProject == "PRJNA418899" | #fastprep/quickgene
           BioProject == "PRJEB2686" | # QIAamp DNA Mini Kit (Qiagen)
           BioProject == "PRJEB2655" | # masterpure
           BioProject == "PRJEB2870" | # QIAamp DNA Mini Kit (Qiagen)
           BioProject == "PRJNA368710" | # DNA Blood and Tissue Kit 69504 Qiagen
           BioProject == "PRJNA369475" | #fastprep/quickgene
           BioProject == "PRJEB11627" #| # MagAttract HMW DNA kit (Qiagen) 
           # BioProject == "PRJEB2812"
           ) %>% # Qiagen QIAcube
  ggplot(aes(x = as.numeric(year),
             color = BioProject,
             y = PCN)) +
  geom_point()
```
this is just over the course of a decade, but it looks like these are all pretty similar ranges between these projects

let's check: do we see signal for extraction technique...?

```{r}
setMC %>%
  mutate(extec = ifelse(BioProject == "PRJEB9575", "qiaxtractor",
                        ifelse(BioProject == "PRJNA418899",
                               "fastprep_quickgene",
                               ifelse(BioProject == "PRJEB2686",
                                      "QIAamp_DNA_Mini",
                                      ifelse(BioProject == "PRJEB2655",
                                             "masterpure",
                                             ifelse(BioProject == "PRJEB2870",
                                                    "QIAamp_DNA_Mini",
                                                    ifelse(BioProject == "PRJNA368710",
                                                           "DNA_Blood&Tissue69504Qiagen",
                                                           ifelse(BioProject == "PRJNA369475",
                                                                  "fastprep_quickgene",
                                                                  ifelse(BioProject == "PRJEB11627",
                                                                         "MagAttractHMWQiagen",
                                                                         ifelse(BioProject == "PRJEB2812",
                                                                                "qiacube",
                                                                                NA)))))))))) %>%
  filter(!is.na(extec)) %>%
  select(sample, Instrument, BioProject, PCN,
         issinglecopy, year, CC, strain:mecA, prepost, centername, extec) %>%
  filter(BioProject != "PRJEB2812") %>%
  ggplot(aes(x = as.numeric(year),
             color = BioProject,
             shape = extec,
             y = PCN)) +
  geom_point()
```
BioProjects have a very narrow time range of collection for the not-old one in this case,
can exclude that from vis

```{r}
setMCex <- setMC %>%
  mutate(extec = ifelse(BioProject == "PRJEB9575", "qiaxtractor",
                        ifelse(BioProject == "PRJNA418899",
                               "fastprep_quickgene",
                               ifelse(BioProject == "PRJEB2686",
                                      "QIAamp_DNA_Mini",
                                      ifelse(BioProject == "PRJEB2655",
                                             "masterpure",
                                             ifelse(BioProject == "PRJEB2870",
                                                    "QIAamp_DNA_Mini",
                                                    ifelse(BioProject == "PRJNA368710",
                                                           "DNA_Blood&Tissue69504Qiagen",
                                                           ifelse(BioProject == "PRJNA369475",
                                                                  "fastprep_quickgene",
                                                                  ifelse(BioProject == "PRJEB11627",
                                                                         "MagAttractHMWQiagen",
                                                                         ifelse(BioProject == "PRJEB2812",
                                                                                "qiacube",
                                                                                NA)))))))))) %>%
  filter(!is.na(extec))
  
setMCex %>%
  select(sample, Instrument, BioProject, PCN,
         issinglecopy, year, CC, strain:mecA, prepost, centername, extec) %>%
  filter(BioProject != "PRJEB2812") %>%
  ggplot(aes(x = PCN,
             fill = as.factor(BioProject))) +
  geom_histogram(position = "stack") +
  facet_wrap(~extec, ncol = 2)
```

what if we look by strain?
```{r}
setMCex %>%
  filter(BioProject != "PRJEB2812") %>%
  ggplot(aes(x = PCN,
             fill = as.factor(strain))) +
  geom_histogram(position = "stack") +
  facet_wrap(~extec, ncol = 2)
```
okay, so the 2 peaks for fastprep are likely generated by STRAIN DIFFERENCES
these have identical protocols and were from the young et al papers 
let's zoom in
```{r}
setMCex %>%
  filter(extec == "fastprep_quickgene") %>%
  ggplot(aes(x = PCN,
             fill = strain)) +
  geom_histogram() +
  facet_wrap(~BioProject, ncol = 1)
```

type the protocols
```{r}
setMCex <- setMCex %>%
  mutate(extype = ifelse(BioProject == "PRJEB9575" |
                           BioProject == "PRJEB2686" |
                           BioProject == "PRJEB2870" |
                           BioProject == "PRJNA368710" |
                           BioProject == "PRJEB11627" |
                           BioProject == "PRJEB2812",
                         "silica",
                         ifelse(BioProject == "PRJNA418899" |
                                  BioProject == "PRJNA369475",
                               "same_unknown", "salt")))
```

```{r}
setMCex %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = extype)) +
  geom_point()
```
looks good!!!

```{r}
setMCex %>%
  filter(as.numeric(year) > 2000) %>%
  ggplot(aes(x = PCN)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~extype, ncol = 1)
```
the difference in pcn between kits might be real
let's check some other things too

```{r}
setMCex %>%
  group_by(Instrument) %>%
  mutate(ins_ct = n()) %>%
  distinct(Instrument, ins_ct)
```
3 different instruments used...
most were sequenced on hiseq 2000

```{r}
setMCex %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = extype,
             shape = Instrument)) +
  geom_point()
```

```{r}
setMCex %>%
  filter(as.numeric(year) > 2000) %>%
  ggplot(aes(x = PCN, fill = extype)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~Instrument, ncol = 1)
```

it looks like instrument should not matter, but extraction type matters a little

let's check library prep maybe....
never mind that's not in the metadata

GET DATE OF SEQUENCING
"pubmet" files on pt181 github

code borrowed from pT181 reunifyG
```{r load sra metadata}
### import all sra downloaded info
pubmet_20150601_20191231 <- read_csv("pubmet-20150601-20191231.txt")
pubmet_20200101_20220530 <- read_csv("pubmet-20200101-20220530.txt")
pubmet_20220601_pres <- read_csv("pubmet-20220601-pres.txt")
pubmet_onbefore20150530 <- read_csv("pubmet-onbefore20150530.txt")

# combine
c1 <- intersect(colnames(pubmet_20150601_20191231), colnames(pubmet_20200101_20220530))
b1 <- rbind(pubmet_20150601_20191231[c1], pubmet_20200101_20220530[c1])

c2 <- intersect(colnames(b1), colnames(pubmet_20220601_pres))
b2 <- rbind(b1[c2], pubmet_20220601_pres[c2])

c3 <- intersect(colnames(b2), colnames(pubmet_onbefore20150530))
b3 <- rbind(b2[c3], pubmet_onbefore20150530[c3])

#clean up the environment
#rm(pubmet_20150601_20191231)
#rm(pubmet_20200101_20220530)
#rm(pubmet_20220601_pres)
#rm(pubmet_onbefore20150530)
```

```{r}
b3 %>%
  select(Experiment,
         Run, BioProject, ReleaseDate, create_date, `ENA-FIRST-PUBLIC (run)`,
         `ENA-LAST-UPDATE (run)`, Collection_Date, INSDC_first_public, 
         INSDC_last_update, ena_first_public, ENA_first_public) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject",
                             "ENA-FIRST-PUBLIC (run)",
                             "ENA-LAST-UPDATE (run)",
                             "Collection_Date")) %>%
  #group_by(Experiment) %>%
  #mutate(count = n()) %>%
  #filter(count > 1)
  select(-Run) %>%
  distinct() %>%
  filter(is.na(`ENA-FIRST-PUBLIC (run)`)) # all have ReleaseDate
# not all have create_date, ENA-FIRST-PUBLIC (run), ENA-LAST-UPDATE (run), etc
```

let's work with release date
```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  ggplot(aes(x = ReleaseDate, y = PCN, color = extype)) +
  geom_point()
```

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  ggplot(aes(x = as.numeric(year), y = ReleaseDate)) +
  geom_hex()
```

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(extype == "silica" &
           Instrument == "Illumina HiSeq 2000") %>%
  ggplot(aes(x = ReleaseDate, y = PCN,
             color = BioProject)) +
  geom_point()
```
There appears to be an effect of BioProject if we ctl for extraction technique & instrument and just plot releasedate on x axis
this is likely explained by year of collection...

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(extype == "silica" &
           Instrument == "Illumina HiSeq 2000") %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             color = BioProject)) +
  geom_point()
```
yeah it is because they were collected earlier it sure seems

let's check strain

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(extype == "silica" &
           Instrument == "Illumina HiSeq 2000") %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             color = strain)) +
  geom_point()
```
IT IS NOT DUE TO STRAIN BACKGROUND!!!!! we can just look at strain 2

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(extype == "silica" &
           Instrument == "Illumina HiSeq 2000" &
           strain == "S99.5_2") %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             color = BioProject,
             shape = centername)) +
  geom_point()
```
ALL SEQUENCED AT SC!!!!!!!

### looking at plots again
```{r year of collection x pcn; color extraction shape instrument}
setMCex %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = extype,
             shape = Instrument)) +
  geom_point()
```

```{r}
setMCex %>%
  filter(year > 1995) %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = extype,
             shape = Instrument)) +
  geom_point()
```

```{r}
setMCex %>%
  filter(year < 1995) %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = extype,
             shape = Instrument)) +
  geom_point()

setMCex %>%
  filter(year < 1995) %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = centername,
             shape = strain)) +
  geom_point()
```

looking at strain 2, sanger center, silica extraction, hiseq 2000 samples
```{r}
setMCex %>%
  filter(strain == "S99.5_2",
         centername == "SC",
         extype == "silica",
         Instrument == "Illumina HiSeq 2000"
       ) %>%
  ggplot(aes(x = as.numeric(year), y = PCN, color = BioProject
             )) +
  geom_point()
```
definitely a decline
run regression to confirm

```{r regression on all same technique strain 2}
strain2_sc_silica_2k <- setMCex %>%
  filter(strain == "S99.5_2",
         centername == "SC",
         extype == "silica",
         Instrument == "Illumina HiSeq 2000"
       )

strain2_sc_silica_2k

s2scsi2k <- lm(PCN ~ as.numeric(year), strain2_sc_silica_2k)
summary(s2scsi2k)
```

```{r min max year for same collection mechanism}
min(strain2_sc_silica_2k$year)
max(strain2_sc_silica_2k$year)
```


```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(extype == "silica" &
           Instrument == "Illumina HiSeq 2000",
         strain == "S99.5_2") %>%
  ggplot(aes(x = ReleaseDate, y = PCN,
             color = BioProject)) +
  geom_point()
```
looks like there's no impact of release date

```{r check for significance in greatly reduced set}
s2silica <- b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(strain == "S99.5_2",
         centername == "SC",
         extype == "silica",
         Instrument == "Illumina HiSeq 2000"
       )

lm.silica.collection <- lm(PCN ~ as.numeric(year)*BioProject, data = s2silica)
lm.silica.release <- lm(PCN ~ ReleaseDate*BioProject, data = s2silica)

summary(lm.silica.collection)
summary(lm.silica.release)

s2silica %>%
  arrange(desc(ReleaseDate
          ))
```
But we CAN'T do control for bioproject because obviously all the old samples are on the same one

let's go back to the 3 bioprojects that cross the 1995 line

### effect of extraction technique
```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(extype == "silica",
         strain == "S99.5_2") %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             color = centername,
             shape = BioProject)) +
  geom_point()
```

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
#  filter(strain == "S99.5_2",
 #        year > 1995) %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             #color = centername,
             color = extype,
             #shape = BioProject
             shape = Instrument)) +
  geom_point()
```

```{r}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(year > 1995) %>%
  ggplot(aes(x = as.numeric(year), y = PCN,
             #color = centername,
             color = extype,
             #shape = BioProject
             shape = Instrument)) +
  geom_point()
```

```{r histogram extraction type pcn by strain}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(year > 1995) %>%
  group_by(strain) %>%
  mutate(straincount = n()) %>%
  filter(straincount > 1) %>%
  ungroup() %>%
  ggplot(aes(x = PCN,
             fill = extype
             )) +
  geom_histogram() +
  facet_wrap(~strain)
```

```{r box plot extraction type pcn by strain}
b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(year > 1995) %>%
  group_by(strain) %>%
  mutate(straincount = n()) %>%
  filter(straincount > 1) %>%
  ungroup() %>%
  ggplot(aes(x = strain,
             y = PCN,
             fill = extype
             )) +
  geom_boxplot() 
```


```{r}
excheck <- b3 %>%
  select(Experiment, Run, BioProject, ReleaseDate) %>%
  right_join(setMCex, by = c("Experiment" = "sample",
                             "BioProject")) %>%
  select(-Run) %>%
  distinct() %>%
  select(Experiment: ReleaseDate, Instrument, year,
         PCN, strain, type, prepost:extype) %>%
  filter(year > 1995) %>%
  group_by(strain) %>%
  mutate(straincount = n()) %>%
  filter(straincount > 1) %>%
  

exT = glmmTMB(PCN ~ as.numeric(year) + extype + strain, data = excheck,
             family = Gamma(link="log"))
summary(exT)
```
don't actually need a glmm for this...
no random effects

```{r regression for strain and extraction technique}
excheckM <- lm(PCN ~ strainFac + extype, data = excheck)
summary(excheckM)
```

```{r}
excheck %>%
  ggplot(aes(x = PCN)) +
  geom_histogram()
```

## investigating S99.5_27 2009-ish peak
```{r}
set2.5backup %>%
  filter(year > 2007 &
           year < 2011) %>%
  filter(strain == "S99.5_27") %>%
  count(BioProject)
```
PRJEB4140 has 1k+ isolates
what are they up to?

```{r}
set2.5backup %>%
  filter(BioProject == "PRJEB4140") %>% # 1083 isolates
  ggplot(aes(x #= as.numeric(year), y 
             = PCN)) +
  #geom_point()
  geom_histogram()
```
all isolates in PRJEB4140 were collected in 2008
They are all single copy

Strain 27 was excluded and rerun in original manuscript

# returning to glmm
exclude strain 2 and rerun

```{r oe5 without strain 2}
oe5no2 <- oe5 %>%
  filter(strain != "S99.5_2")
```

```{r glmm for time and strain without strain 2}
m22 = glmmTMB(PCN ~ numYear + strainFac +
                (1|strainFac:BioProject) +
                (1|numYear:BioProject), # AIC is lower with them both included 
        data = oe5no2,
             family = Gamma(link="log"))
summary(m22)
```
relationship with numYear still significant

what does the distribution of years look like in this new dataset
```{r}
oe5no2 %>%
  ggplot(aes(x = numYear, y = PCN, color = strainFac)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r year range for non-s2 oe5}
summary(oe5no2$numYear)
```
without strain 2 and reducing to only bioprojects with 5+ samples
year range is small, 2003-2017, mostly sampled 2015+

the relationship is still significant and negative, even looking at this small range
just not as significant and negative
