---
title: "pT181-reunify"
author: "Megan Phillips"
date: "5/22/2025"
output: html_document
---
# Getting started

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load all packages}
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(forcats)
library(gggenes)
library(ez)
library(DescTools)
library(ggpubfigs)
library(ggpubr)
library(FSA)
library(ggstream)
```

```{r load sra metadata}
### import all sra downloaded info
pubmet_20150601_20191231 <- read_csv("pubmet-20150601-20191231.txt")
pubmet_20200101_20220530 <- read_csv("pubmet-20200101-20220530.txt")
pubmet_20220601_pres <- read_csv("pubmet-20220601-pres.txt")
pubmet_onbefore20150530 <- read_csv("pubmet-onbefore20150530.txt")

# combine
c1 <- intersect(colnames(pubmet_20150601_20191231), colnames(pubmet_20200101_20220530))
b1 <- rbind(pubmet_20150601_20191231[c1], pubmet_20200101_20220530[c1])

c2 <- intersect(colnames(b1), colnames(pubmet_20220601_pres))
b2 <- rbind(b1[c2], pubmet_20220601_pres[c2])

c3 <- intersect(colnames(b2), colnames(pubmet_onbefore20150530))
b3 <- rbind(b2[c3], pubmet_onbefore20150530[c3])

#clean up the environment
rm(pubmet_20150601_20191231)
rm(pubmet_20200101_20220530)
rm(pubmet_20220601_pres)
rm(pubmet_onbefore20150530)
```

```{r load staphopia v2 report}
staphopia_v2_all_report <- read_delim("staphopia_v2_all_report.txt", 
                                      delim = "\t", escape_double = FALSE, 
                                      trim_ws = TRUE)
```

```{r create datasets eg all86k}
# reduce number of columns
br3 <- b3 %>%
  select(Run:Collection_Date)

staphopia_v2_all_red <- staphopia_v2_all_report %>%
  select(sample:estimated_genome_size, mlst_blast_st:mlst_ariba_yqil)

all86k <- left_join(staphopia_v2_all_red, br3,
                      by = c("sample" = "Experiment"))
```

```{r 83k confirmation}
all86k %>%
  distinct(sample) # 83,366 unique experiments in the SA dataset
```

loading sccmec data for s.aureus from allthebacteria
```{r load sccmec data}
allthebacteria_sccmec <- read_delim("allthebacteria-sccmec.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

join sccmec with the staphopia set 
```{r add sccmec to all86k}
all86k <- left_join(all86k, allthebacteria_sccmec, by = c("BioSample" = "sample")) 
```

Loading in ST & CC information
```{r load st & cc}
# which STs are in which CCs
stcc <- read_delim("st-cc.tsv", 
    delim = ",", #escape_double = FALSE, 
    trim_ws = TRUE)

stcc$ST <- str_replace(stcc$ST, "ST", "")
```

```{r add stcc to all86k}
all86k <- left_join(all86k, stcc,
                    by = c("mlst_blast_st" = "ST"))
```

```{r how many st & cc?}
all86k$mlst_blast_st %>%
  fct_lump_n(10) %>%
  table()
#gplot(aes(x = mlst_blast_st))

all86k$CC %>%
  fct_lump_n(10) %>%
  table()
```

import Vishnu's strain groups
```{r import strain}
vstraingroup <- read_delim("isolate_substrain_strain_freq_st_CC.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

adding strain group to all86k df
```{r add strain to all86k}
all86k <- left_join(all86k, vstraingroup,
          by = c("sample" = "isolate"))

all86k %>%
  filter(!is.na(strain)) %>%
  distinct(sample) #56,770
```

let's get the 337 complete genomes that start with dnaA
```{r load dnaA-1st complete genomes}
dnaA_start_coords <- read_delim("fromV2git/timsfigsData/dnaA_start_coords.gfs", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)

#info on the 14 integrated complete dnaA-1st
X14_description_ncbi <- read_delim("14.description.ncbi.txt", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

```{r load complete over threshold}
completeover <- read_csv("completegenomes_samplesoverorequal4400.txt", 
    col_names = FALSE)

completeover <- completeover %>%
  rename("sseqid" = "X1")

completeblast <- read_delim("completegenomes.blast.1.out", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)

completeblast <- completeblast %>%
  rename("qseqid" = "X1",
         "sseqid" = "X2",
         "pident" = "X3",
         "length" = "X4",
         "mismatch" = "X5",
         "gapopen" = "X6",
         "qstart" = "X7",
         "qend" = "X8",
         "sstart" = "X9",
         "send" = "X10",
         "evalue" = "X11",
         "bitscore" = "X12")
```

load blast hit distribution
```{r load hit lengths}
totallength <- read_table2("samples-blasthit-totallength.txt")
```

```{r Load c25 color pal}
c25 <- c(
  "dodgerblue2",
  "#E31A1C", # red
  "green4",
  "#6A3D9A", # purple
  "#FF7F00", # orange
  "black",
  "gold1",
  "skyblue2",
  "#FB9A99", # lt pink
  "palegreen2",
  "#CAB2D6", # lt purple
  "#FDBF6F", # lt orange
  "gray70",
  "khaki2",
  "maroon",
  "orchid1",
  "deeppink1",
  "blue1",
  "steelblue4",
  "darkturquoise",
  "green1",
  "yellow4",
  "yellow3",
  "darkorange4",
  "brown"
)

barplot(rep(1, length(c25)), col = c25, names.arg = 1:length(c25))
```

```{r strain colors}
strain.c25 <- c(
  "S99.5_10" = "dodgerblue2", #1
  "S99.5_11" = "palegreen2",
  "S99.5_13" = "maroon",
  "S99.5_15" = "orchid1",
  "S99.5_16" = "deeppink1",
  "S99.5_17" = "blue1",
  "S99.5_18" = "steelblue4",
  "S99.5_2" = "#E31A1C", # red, 2
  "S99.5_20" = "darkturquoise",
  "S99.5_23" = "green4", #3
  "S99.5_24" = "darkorange4",
  "S99.5_26" = "brown",
  "S99.5_27" = "#6A3D9A", # purple, 4
  "S99.5_4" = "#CAB2D6", # lt purple
  "S99.5_45" = "green1",
  "S99.5_49" = "#FB9A99", # lt pink
  "S99.5_5" = "#FF7F00", # orange, 5
  "S99.5_7" = "gold1",
  "S99.5_8" = "skyblue2",
  "Other" = "gray70"
)

allele.pallete <- c(
  "S99.5_10" = "dodgerblue2", #1
  "S99.5_11" = "palegreen2",
  "S99.5_13" = "maroon",
  "S99.5_15" = "orchid1",
  "S99.5_16" = "deeppink1",
  "S99.5_17" = "blue1",
  "S99.5_18" = "steelblue4",
  "S99.5_2" = "#E31A1C", # red, 2
  "S99.5_20" = "darkturquoise",
  "S99.5_23" = "green4", #3
  "S99.5_24" = "darkorange4",
  "S99.5_26" = "brown",
  "S99.5_4" = "#CAB2D6", # lt purple
  "S99.5_45" = "green1",
  "S99.5_49" = "#FB9A99",
  "S99.5_5" = "#FF7F00", # orange, 5
  "S99.5_7" = "gold1",
  "S99.5_8" = "skyblue2"
)

top7strain2.5 <- c(
  "S99.5_10" = "dodgerblue2", #1
  "S99.5_2" = "#E31A1C", # red, 2
  "S99.5_23" = "green4", #3
  "S99.5_27" = "#6A3D9A", # purple, 4
  "S99.5_5" = "#FF7F00", # orange, 5
  #"black",
  "S99.5_7" = "gold1",
  "S99.5_8" = "skyblue2",
  "Other" = "gray70"
)

barplot(rep(1, length(strain.c25)), col = strain.c25, names.arg = 1:length(strain.c25)
        )
```


```{r backup}
set2.5backup <- read_delim("set2.5_20250324_safety.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)
```

# quick check: median pcn by time 
```{r}
set2.5backup %>%
  filter(issinglecopy == F) %>%
  select(PCN, prepost) %>%
  group_by(prepost) %>%
  mutate(mean = mean(PCN),
         med = median(PCN)) %>%
  select(prepost, mean, med) %>%
  distinct()

(14.53175 * 4440)/2800000 # post median %
(55.68119 * 4440)/2800000 # post median %
```


# Spread across the species

## process the dates
Doing some date fixing
```{r process Collection_Date metadata}
sampdate_dateEntered <- all86k %>% 
  filter(!is.na(Collection_Date), Collection_Date != "Missing",
         Collection_Date != "missing",
         Collection_Date != "NA",
         Collection_Date != "not collected",
         Collection_Date != "not available",
         Collection_Date != "Not applicable",
         Collection_Date != "not provided",
         Collection_Date != "not applicable",
         Collection_Date != "Not collected",
         Collection_Date != "not determined",
         Collection_Date != "Not Applicable",
         Collection_Date != "Not available") %>%
  distinct() 

sampdate_dateUsable1 <- sampdate_dateEntered %>%
  filter (Collection_Date != "1960/1970",
         Collection_Date != "05/06/2010",
         Collection_Date != "07-12-11",
         Collection_Date != "2014-01/2015-08",
         Collection_Date != "17-04-12",
         Collection_Date != "19-04-12",
         Collection_Date != "2014-11/2016-01",
         Collection_Date != "1998/2012",
         Collection_Date != "2006/2007",
         Collection_Date != "0000",
         Collection_Date != "1980's",
         Collection_Date != "09.04.2010",
         Collection_Date != "1994-09/2003-10",
         Collection_Date != "2015-17")

rm(sampdate_dateEntered)

sampdate_dateUsable1$new_Date <- sampdate_dateUsable1$Collection_Date %>%
  str_replace("2016-0.-../2016-1.*$", "2016") %>%
  str_replace("2015-0.-../2015-1.-..", "2015") %>%
  str_replace("2005-0.-../2005-1.-..", "2005") %>%
  str_replace("2004-0.-../2004-1.-..", "2004") %>%
  str_replace("2016-0./2016-1.", "2016") %>%
  str_replace("2010-0.-../2010-1.-..", "2010") %>%
  str_replace("^../../..$", "notspp") %>%
  str_replace("^.././..$", "notspp") %>%
  str_replace("^./../..$", "notspp") %>%
  str_replace("^..-..-..$", "notspp") %>%
  str_replace("^..../....$", "notspp") %>%
  str_replace("^../../....$", "notspp") %>%
  str_replace("pre.*$", "notspp") %>%
  str_replace("Pre.*$", "notspp") %>%
  str_replace("^././..$", "notspp") %>%
  str_replace("Jan", "01") %>%
  str_replace("Feb", "02") %>%
  str_replace("Mar", "03") %>%
  str_replace("Apr", "04") %>%
  str_replace("May", "05") %>%
  str_replace("Jun", "06") %>%
  str_replace("Jul", "07") %>%
  str_replace("Aug", "08") %>%
  str_replace("Sep", "09") %>%
  str_replace("Oct", "10") %>%
  str_replace("Nov", "11") %>%
  str_replace("Dec", "12") %>%
  na_if("notspp")

sampdate_dateUsable2 <- sampdate_dateUsable1 %>%
  filter(!is.na(new_Date))

rm(sampdate_dateUsable1)

sampdate_dateUsable2$year <- sampdate_dateUsable2$new_Date %>%
  str_replace("^..-..-","") %>%
  str_replace("-..-..$","") %>%
  str_replace("^..-.. ", "") %>%
  str_replace("^..-", "") %>%
  str_replace("-..$", "")
```

```{r add good dates to all86k}
all86k <- left_join(all86k, sampdate_dateUsable2)
rm(sampdate_dateUsable2)

all86k$prepost <- with(all86k,ifelse(as.numeric(all86k$year) < 1995, "pre", "post"))
```

Load in the contigs that have pt181
```{r}
contiglist <- read_csv("20241209-contiglist.csv", 
    col_names = FALSE)

contiglist <- contiglist %>%
  rename("sample" = X1,
         "pT181contigSA" = X2)

# add contig list for pT181 to all86k df
all86k <- left_join(all86k, contiglist, by = "sample")
```

```{r checking date stats}
all86k %>%
  filter(!is.na(year)) %>%
  distinct(sample) # 42,522 genomes are dated

all86k$year %>%
  as.numeric() %>%
  summary() # earliest = 1884, latest = 2022

all86k$prepost %>%
  table()
```

## over strains & over time


# Mutation analysis

## snippy output

Load snp position data
```{r load snippy output}
# from stragglers
# run on 86 stragglers = any that had pT181 (incl. undated + not rank-checked)
stragglersnps <- read_table("allstragglersnps.tab")
stragglersnps$isSA <- T

# from original set
# everything with hits covering over 4400bp
snps <- read_delim("allsnps.tsv", 
                      delim = "\t", escape_double = FALSE, 
                      trim_ws = TRUE)
snps$isSA <- T

# from non-SA which exceeded hit length requirements
nonSAsnps <- read_table("nonSA63sampsnps.tab")
nonSAsnps$isSA <- F

```

Make one big df
```{r combine mut info}
df1 <- full_join(nonSAsnps, stragglersnps)
df2 <- full_join(df1, snps) %>%
  select(SAMPLE, POS, TYPE, REF, ALT, isSA)
df2$Experiment <- df2$SAMPLE %>%
  str_remove("-cont.*.$")
```


### binning identical samples
```{r}
js <- left_join(set2.5backup, df2, by = c("sample" = "Experiment")) %>%
  select(sample, strain, PCN, year, issinglecopy, prepost, POS:ALT)
js$allele <- paste0(js$POS, "-", js$TYPE, "-", js$REF, "-", js$ALT)

jss <- js %>%
  group_by(sample, PCN) %>%
  summarise(plasAllele = paste(allele, collapse = "-"), .groups = "drop") 

jss %>%
  select(plasAllele) %>%
  distinct() # 229 unique plasmid sequences
```

```{r Fig. 8: plotting alleles x PCN}
left_join(jss, js) %>%
  select(sample, PCN, strain, plasAllele, year, prepost) %>%
  distinct() %>%
  group_by(plasAllele) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  filter(count > 2) %>%
  filter(PCN > 2) %>%
  filter(prepost == "pre") %>%
  ggplot(aes(x = plasAllele, y = PCN, color = strain)) +
  #geom_violin() +
  geom_point() +
  scale_y_sqrt()

# there's not just one allele in the early samples generating the high PCN

### fig. 8 plot
left_join(jss, js) %>%
  select(sample, PCN, strain, plasAllele, year, prepost) %>%
  distinct() %>%
  group_by(plasAllele) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  filter(count > 2) %>%
  filter(PCN > 2) %>%
  ggplot(aes(x = forcats::fct_reorder(plasAllele, PCN, .fun = median), y = PCN, color = strain)) +
  geom_point() +
  scale_y_sqrt() +
  theme_bw() +
  theme(axis.text.x = element_blank(),
  axis.ticks = element_blank()) +
  scale_color_manual(values = allele.pallete) +
  xlab("Plasmid Allele")

left_join(jss, js) %>%
  select(sample, PCN, strain, plasAllele, year, prepost) %>%
  distinct() %>%
  group_by(plasAllele) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  filter(count > 2) %>%
  filter(PCN > 2) %>%
  select(strain) %>%
  unique()

# 18 strains -- did not bin for "Other"
# make "otherless + no 27" palette
```

```{r identical plasmids: single & multicopy}
left_join(jss, js) %>%
  select(sample, PCN, plasAllele, issinglecopy) %>%
  distinct() %>%
  ggplot(aes(x = plasAllele, fill = issinglecopy)) +
  geom_bar(position = "stack") +
  scale_y_sqrt()
```
There are identical plasmids that are both single and multicopy

are there mutations uniue to SA/nonSA isolates?
```{r SA exclusive?}
# S. aureus exclusive, occuring more than once
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  group_by(POS, TYPE, ALT, isSA) %>%
  mutate(count = n()) %>%
  distinct(POS, TYPE, ALT, isSA, count) %>%
  pivot_wider(names_from = isSA,
              values_from = count) %>%
  filter(is.na(`FALSE`)) %>%
  filter(`TRUE` >= 2) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()

# non-S. aureus exclusive, occuring more than once
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  group_by(POS, TYPE, ALT, isSA) %>%
  mutate(count = n()) %>%
  distinct(POS, TYPE, ALT, isSA, count) %>%
  pivot_wider(names_from = isSA,
              values_from = count) %>%
  filter(is.na(`TRUE`)) %>%
  filter(`FALSE` >= 2) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```

```{r}
df2 %>%
  filter(isSA == F) %>% # nonSA only
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  filter(POS <= 794 | # filtering for CDS only
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049)) %>%
  filter(TYPE == "snp")
```



```{r prepping to filter 4190 spike to create ref plot}
POS <- 4190
TYPE <- "snp"
REF <- "G"
ALT <- "T"

dfForAntijoin <- data.frame(POS = POS, TYPE = TYPE, REF = REF, ALT = ALT)
```

```{r df relative to consensus}
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  filter(POS == 4190 & ALT == "T" & TYPE == "snp") %>% # 7,706 rows have this allele
  select(POS, ALT, TYPE) %>%
  table()

dfA <- anti_join(df2, dfForAntijoin, by = c("POS", "TYPE", "REF", "ALT")) %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") #20,505 rows -- correct

# 7,988 - 7,706 = 282
dfA4190 <- data.frame(
  SAMPLE = rep("sampleX", 282),
  POS = rep(4190, 282),
  TYPE = rep("snp", 282),
  REF = rep("G", 282),
  ALT = rep("G", 282)
)

presplot <- full_join(dfA, dfA4190)
```

```{r quick math}
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") #28,211 rows

28211 - 20505
```

```{r snippy plot}
# this is relative to consensus
presplot %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram(binwidth = 10) +
  scale_fill_manual(values = c25) +
  theme_bw() +
  labs(x = "Plasmid position")
```

```{r missing anything in presplot at high frequency?}
presplot %>%
  select(POS, TYPE, ALT) %>%
  group_by(POS, TYPE, ALT) %>%
  mutate(count = n()) %>%
  distinct() %>%
  filter(count > 3000)

presplot %>%
  select(POS, TYPE, ALT) %>%
  group_by(POS, TYPE, ALT) %>%
  mutate(count = n()) %>%
  distinct() %>%
  filter(POS == 2464)
```


```{r plotting mutations}
# mutation frequency relative to edited
# df2 is relative to reference, had to edit to deal with spike at 4190
presplot %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram(binwidth = 1)

# allele count
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>% 
  group_by(POS) %>% 
  summarise(NumberOfVar = n_distinct(TYPE, ALT)) %>%
  ggplot(aes(x = as.numeric(POS), y = NumberOfVar)) +
  geom_col()
```

that spike in pre also might be high freq
let's check on it
```{r checking pre spike}
presplot %>%
  filter(POS > 2200 & POS < 2500) %>%
  filter(TYPE == "complex") %>%
  select(POS, ALT) %>%
  table()

3064 / 7988
```
Highest freq mut is in 3064 samples
Still not the majority of samples!

```{r alleles plot}
df3 <- df2 %>% 
  group_by(POS) %>% 
  summarise(NumberOfVar = n_distinct(ALT))

df3 %>%
  mutate(Alleles = NumberOfVar + 1) %>%
  ggplot(aes(x = POS, y = Alleles)) +
  geom_col(width = 1) +
  xlab("Plasmid position") +
  theme_bw()
```

```{r combining Fig. 2A & B}
df2 %>%
  select(POS, TYPE, ALT) %>% # for all samples with mut, what is allele & type at each site?
  distinct() %>% # distinct mutations (we don't need to know sample count) -- 680 rows
  group_by(POS, TYPE) %>%
  mutate(countPT = n()) %>% # @ all sites how many different mutations of each type?
  select(-ALT) %>% # don't count alts, we know how many now
  distinct() %>% 
  ggplot(aes(x = as.numeric(POS), y = countPT, #color = TYPE, 
             fill = TYPE)) +
  geom_col(width = 10,
           position = "stack") +
  #geom_bar(width = 1) +
  theme_bw() +
  labs(y = "# Alt Alleles",
       x = "Plasmid Position",
       fill = "Mutation type")

# think about how we want to represent this / the axes. what is this?
# need to learn or establish the binwidth / width for these!!!!
```

```{r Fig. 2 with 10bp binwidth}
df2 %>%
  select(POS, TYPE, ALT) %>% # for all samples with mut, what is allele & type at each site?
  distinct() %>% # distinct mutations (we don't need to know sample count) -- 680 rows
  mutate(POSn = str_replace(POS, "\\d(?=\\D*$)", "0"))%>%
  select(POSn, TYPE, ALT) %>%
  group_by(POSn, TYPE) %>%
  mutate(countTYPEPOSn = n()) %>%
  distinct() %>%
  ggplot(aes(x = as.numeric(POSn), #y = countPT, #color = TYPE, 
             fill = TYPE)) +
  geom_bar(position = "stack") +
  theme_bw() +
  labs(y = "# Alt Alleles",
       x = "Plasmid Position (10bp binwidth)",
       fill = "Mutation type")
```
ALSO need a plot with the genes



Let's look at the gene plot
```{r Fig. 2 gene plot}
baktaRef1 <- data.frame(
  Type = c("cds", "sorf", "cds", "cds", "oriC", "cds", "ncRNA", "cds"),
  Start = c(0, 862, 956, 2521, 3763, 3948, 4100, 4289),
  Stop = c(794, 912, 2335, 3762, 4288, 4049, 4194, 4439)
)

baktaRef2 <- data.frame(
  Type = c("repC", "sorf", "tetK", "pre", "oriC", "cds", "ncRNA", "repC"),
  Start = c(0, 862, 956, 2521, 3763, 3948, 4100, 4289),
  Stop = c(794, 912, 2335, 3762, 4288, 4049, 4194, 4439)
)

baktaRef2$length <- baktaRef2$Stop - baktaRef2$Start

geneplot <- ggplot(baktaRef2, aes(xmin = Start, xmax = Stop, y = "pT181",
                                 fill = Type)) +
  geom_gene_arrow() +
  #geom_gene_label() +
  #facet_wrap(~ molecule, scales = "free", ncol = 1) +
  scale_fill_brewer(palette = "Set3") +
  theme(legend.position="none")

geneplot + theme_bw()
```

## checking on 1954 isolate (& others)

Does the 1954 isolate have core SNPs (relative to reference plasmid)?
```{r}
left_join(set2.5backup, df2, by = c("sample" = "Experiment")) %>%
  filter(as.numeric(year) < 1960) %>% # just SRX864932 from 1954
  filter(POS <= 794 | # filtering for CDS only
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049))
```
1954 isolate has NO CDS snps relative to reference


## CDS mutations reworked

We now know that there are significant insertions/deletions occurring in the plasmid, even in coding regions!

We need to revisit the output to estimate a mutation rate, or more accurately, snp accumulation rate in coding regions (CDS)

```{r making snp datasets}
snpSet <- left_join(set2.5backup, df2, by = c("sample" = "Experiment")) %>% 
  # no need to filter cloning vectors since we are just going with HQ SA samples
  filter(POS <= 794 | # filtering for CDS only
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049)) %>%
  filter(TYPE == "snp") %>% # filtering for SNPs only
  group_by(sample) %>%
  mutate(SNPcount = n()) %>%
  select(sample, SNPcount)

cdsSNP <- left_join(set2.5backup, snpSet) %>%
  replace_na(list(SNPcount = 0)) %>%
  select(sample, year, SNPcount, PCN, issinglecopy, strain, type) %>%
  distinct()
```

Recreating figures using new metrics (SNPs, instead of all mutation types)
```{r Fig. 3: snp figs}
ggplot(data = cdsSNP, aes(x = as.numeric(year), y = SNPcount)) +
  #geom_point() +
  geom_jitter(width = 0.3, height = 0.05) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_bw() +
  labs(x = "Year of Collection",
       y = "CDS SNP count")

cdsSNP$SNPcountFct <- factor(cdsSNP$SNPcount, levels=c('4', '3', '2', '1', '0'))

cdsSNP %>%
  mutate(decade = str_replace(year, ".$", "0s")) %>%
  filter(year > 1954) %>%
  ggplot(aes(x = decade, fill = SNPcountFct)) +
  geom_bar(position = "fill") +
  labs(x = "Decade",
       y = "Proportion of Samples",
       fill = "Coding region\nSNP count") +
  theme_bw() +
  scale_fill_manual(values = c25)
```

### how many sites have more than 1 allele (aka any snippy output)?
```{r}
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  select(POS) %>%
  distinct()
# 519 sites have a mutation

df2 %>%
  filter(SAMPLE !="AY672108.1" &
          SAMPLE !="JX893516.1") %>%
  select(ALT, TYPE, POS) %>%
  distinct()
# 664 different mutations happen across the plasmid
```

```{r}
df2 %>%
  filter(SAMPLE !="AY672108.1" &
          SAMPLE !="JX893516.1") %>%
  select(ALT, TYPE, POS) %>%
  group_by(ALT, TYPE, POS) %>%
  mutate(occurance = n()) %>%
  distinct() %>% # 664 -- good
  arrange(desc(occurance))
  #ggplot(aes(x = occurance)) +
  #geom_point()
```
Remember -- 4190 snp T at a really high freq, actually more common allele

<5% of samples == <400 isolates
```{r}
df2 %>%
  filter(SAMPLE !="AY672108.1" &
          SAMPLE !="JX893516.1") %>%
  select(ALT, TYPE, POS) %>%
  group_by(ALT, TYPE, POS) %>%
  mutate(occurance = n()) %>%
  distinct() %>%
  filter(occurance < 400) #  655 rows # add 1 for the 4190 position -> 656
```


##### reflection on CDS mutations vs SNPs
The reference plasmid which we mapped to is 4439 bp long
this is actually just a little short - I expect especially to see insertions, but I suppose deletions are possible

just want to check what kind of stuff is happening in the CDS
```{r cds mut table for 2.5k set}
left_join(set2.5backup, df2, by = c("sample" = "Experiment")) %>%
  filter(POS <= 794 | 
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049)) %>%
  select(TYPE) %>%
  table()
```



##### Mutation accumulation -- look at other species
```{r nonSA mutations in CDS}
df2 %>%
  filter(isSA == F) %>%
  #distinct(SAMPLE) %>% # 60 non SA samples here (that have mut) (at this point - check for cloning vectors)
  filter(POS <= 794 | 
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049)) %>%
  #distinct(SAMPLE) # 23 rows: most nonSA are identical in CDS to SA pT181
  group_by(SAMPLE) %>%
  mutate(mutcount = n()) %>%
  select(SAMPLE, mutcount) %>%
  distinct() 
  
 # ggplot(aes(x = POS, fill = ALT)) +
  #geom_histogram() +
  #theme(legend.position="none")
df2 %>%
  filter(isSA == F) %>%
  filter(SAMPLE == "CP128253.1") %>%
  filter(POS <= 794 | 
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049))
```
CP128253.1 - mutcount 12 - Staphylococcus pseudintermedius strain 20VMG1536 chromosome
This is the only non-cloning vector that is farther than the SA

AY672108.1 & JX893516.1 are cloning vectors

CP092479.1 - mutcount 4 - Staphylococcus haemolyticus strain CGMH-SH51 plasmid unnamed, complete sequence, Staphylococcus haemolyticus,7633,7901,100%,0.0,99.44,33938,
4 is still within the SA set
Everthing else is <=3

are there snps uniue to SA/nonSA isolates?
```{r exclusive snps?}
# S. aureus exclusive, occuring more than once
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  group_by(POS, TYPE, ALT, isSA) %>%
  mutate(count = n()) %>%
  distinct(POS, TYPE, ALT, isSA, count) %>%
  pivot_wider(names_from = isSA,
              values_from = count) %>%
  filter(is.na(`FALSE`)) %>%
  filter(`TRUE` >= 2) #%>%
  #ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  #geom_histogram()

# non-S. aureus exclusive, occuring more than once
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  group_by(POS, TYPE, ALT, isSA) %>%
  mutate(count = n()) %>%
  distinct(POS, TYPE, ALT, isSA, count) %>%
  pivot_wider(names_from = isSA,
              values_from = count) %>%
  filter(is.na(`TRUE`)) %>%
  filter(`FALSE` >= 2)# %>%
  #ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  #geom_histogram()
```
There are mutations which are "private" to SA/non-SA --- most of them are private to SA, which makes sense given the composition of the dataset
in Non-sa samples, there is 1 snp in 2 isolates and 1 complex mut in 3 isolates

```{r}
df2 %>%
  filter(SAMPLE !="AY672108.1" & # filtering out the cloning vectors
          SAMPLE !="JX893516.1") %>%
  filter(POS == 1943 | POS == 3359)
```
All different experiments for the snp/complex mutation havers


### back to the regressions on snps
Let's rerun the regressions for strains, time, etc
```{r snp regression}
#glm(SNPcount ~ as.numeric(year) + as.factor(strain),
 #   data = cdsSNP, family = poisson) %>%
  #summary()

lm(SNPcount ~ as.numeric(year),
   data = cdsSNP) %>%
  summary()
```
Date of collection has a significant impact on SNPs accumulated in CDS

```{r size of cds}
cdslen <- (4439 - 4323) + (4049 - 3948) + (3762 - 2521) + (2335 - 956) + (912 - 862) + (794)
```

```{r}
# 0.004234 cds snps / year in pT181
0.004234 / 3681
```
1.150231e-06 snps/bp/year --> 1.15 SNP/Mbp/year 
this is on the lower end compared to other estimates, but nothing too out-there


Now that we are looking at just the SNPs, more strains jump out as having different amounts!
Let's plot, looking at strain, SC vs not, etc

```{r}
cdsSNP %>%
  ggplot(aes(x = as.numeric(year), y = SNPcount, color = issinglecopy)) +
  geom_point() +
  facet_wrap(~strain)
```
To make it easier, just look at strains with at least 10 plasmids

```{r}
cdsSNP %>%
  group_by(strain) %>%
  mutate(strainCount = n()) %>%
  filter(strainCount >= 10) %>%
  ggplot(aes(x = as.numeric(year), y = SNPcount, color = issinglecopy)) +
  geom_point() +
  facet_wrap(~strain)
```

now re-do the regressions with the 10+ plasmids strains

Can't pipe to glm :(
Make a new df

```{r}
strainCountMin10 <- cdsSNP %>%
  group_by(strain) %>%
  mutate(strainCount = n()) %>%
  filter(strainCount >= 10)

glm(SNPcount ~ as.numeric(year) + as.factor(strain),
    data = strainCountMin10,
    family = poisson) %>%
  summary()

lm(SNPcount ~ as.numeric(year),
   data = strainCountMin10) %>%
  summary()
```
Okay, looks like strain does really have an effect on snps accumulated....

Hmmmmmmmmmmmm maybe look at where these are happening? Where are they localized?

Of course, within a strain especially they are not independent

```{r plotting cds snps by strain}
left_join(set2.5backup, df2, by = c("sample" = "Experiment")) %>%
  filter(TYPE == "snp") %>%
  filter(POS <= 794 | # filtering for CDS only
           POS >= 4323 |
           (POS >= 862 & POS <= 912) |
           (POS >= 956 & POS <= 2335) |
           (POS >= 2521 & POS <= 3762) |
           (POS >= 3948 & POS <= 4049)) %>% 
  ggplot(aes(x = POS, fill = ALT)) +
  geom_histogram() +
  facet_wrap(~strain, scales = "free")
```

Would be tough to actually cluster plasmids by CDS snps -- there is 1 main genotype across all samples



# Plasmid copy number (PCN) calculations
loading in stuff that has been averaged across contig for all strains
```{r load depths and gene positions}
# avg depth for each contig for non-straggler samples
avgdepths <- read_table2("20231222/cleaned_avgdepths.txt", col_names = FALSE)
avgdepths <- avgdepths %>% rename("Experiment" = "X1", "contig" = "X2",
                                  "avgDepth" = "X3")

#straggler depths here
stragglercleaned <- read_table2("straggler.cleaned.avgdepth", 
    col_names = FALSE)
stragglercleaned <- stragglercleaned %>% rename("Experiment" = "X1", "contig" = "X2",
                                  "avgDepth" = "X3")

#lawful+tetK contigs
#tetK = g00000
geneID <- read_table2("20231222/geneID-forR.txt", 
    col_names = FALSE)
geneID <- geneID %>% rename("gene" = "X1", "Experiment" = "X2",
                            "contig" = "X3")

#straggler IDs here
stragger_lawfulpos <- read_table2("stragger.lawfulpos.tsv", 
    col_names = FALSE)
stragger_lawfulpos <- stragger_lawfulpos %>% rename("gene" = "X1",
                                                    "Experiment" = "X2",
                                                    "contig" = "X3")
```

report join to get the gold
```{r get the gold}
avgdepthsgold <- left_join(avgdepths, staphopia_v2_all_report,
                       by = c("Experiment" = "sample")) %>%
  select(Experiment, contig, avgDepth, rank, mlst_blast_st:mlst_ariba_yqil) %>%
  filter(rank == "gold")

stragglercleanedgold <- left_join(stragglercleaned, staphopia_v2_all_report,
                       by = c("Experiment" = "sample")) %>%
  select(Experiment, contig, avgDepth, rank, mlst_blast_st:mlst_ariba_yqil) %>%
  filter(rank == "gold")
```

adding gene position info
```{r add gene location}
geneID$loc <- with(geneID, ifelse(gene=="g00000", 'pT181', 'chrom'))
stragger_lawfulpos$loc <- with(stragger_lawfulpos,
                               ifelse(gene=="g00000", 'pT181', 'chrom'))
```

put the depths together
```{r combine depths}
regdepths <- left_join(avgdepthsgold, geneID,
                        by = c("Experiment", "contig")) %>%
  select(-gene) %>%
  distinct()

strdepths <- left_join(stragglercleanedgold, stragger_lawfulpos,
                        by = c("Experiment", "contig")) %>%
  select(-gene) %>%
  distinct()

alldepths1 <- full_join(regdepths, strdepths)
```


Calculating plasmid copy number (PCN) from relative read depth
```{r calculate PCN}
# get avg for each location for each sample
ratio1 <- alldepths1 %>% 
  group_by(Experiment, loc) %>% 
  summarise(meandepth = mean(avgDepth))

untidy1 <- pivot_wider(ratio1, names_from = loc,
                      values_from = meandepth)

# adding ratio of plasmid to chromosome
untidy1$PCN <- (untidy1$pT181 / untidy1$chrom)

summary(untidy1$PCN)
```

```{r visualize PCN & get nonNA df}
# all PCN
pcnA <- ggplot(data = untidy1, mapping = aes(x = PCN)) +
  geom_histogram(binwidth = 1) +
  theme_bw()

pcn.notNA1 <- untidy1 %>%
  filter(!is.na(PCN))

# NAs are due to no good-enough blast hit for pT181
# due to straggler filtering -- filtering is happening now

mean(pcn.notNA1$PCN)
median(pcn.notNA1$PCN)
sd(pcn.notNA1$PCN)
summary(pcn.notNA1$PCN)

# PCN =< 50
pcnB <- pcn.notNA1 %>%
  filter(PCN <= 50) %>%
  ggplot(mapping = aes(x = PCN)) +
  geom_histogram(binwidth = 0.1) +
  theme_bw()

# PCN =< 3
pcnC <- pcn.notNA1 %>%
  filter(PCN <= 3) %>%
  ggplot(mapping = aes(x = PCN)) +
  geom_histogram(binwidth = 0.01) +
  theme_bw()

cowplot::plot_grid(pcnA, pcnB, pcnC, ncol = 3)
```

```{r Fig. 4 reworking pcn visualization}
# all PCN
pcnAb <- ggplot(data = pcn.notNA1, mapping = aes(x = PCN)) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  scale_x_sqrt()

# PCN =< 3
pcnCb <- pcn.notNA1 %>%
  filter(PCN <= 3) %>%
  ggplot(mapping = aes(x = PCN)) +
  geom_histogram(binwidth = 0.01) +
  theme_bw()

cowplot::plot_grid(pcnAb, pcnCb, ncol = 2)
```

get non-single copy stats
```{r MC stats}
pcn.notNA1 %>%
  filter(PCN > 2) %>%
  summary()
```

Add PCN information to all86k
```{r all86k <- + PCN}
all86k <- left_join(all86k, pcn.notNA1, by = c("sample" = "Experiment")) 
```

This is the set2.5, with 2,460 isolates
```{r checking on set2.5}
all86k %>%
  filter(!is.na(mecA)) %>% # is def SA & we have sccmec
  filter(!is.na(PCN)) %>% # gold, SA, and has pt181
  filter(!is.na(year)) %>% # valid date, is SA
  filter(!is.na(CC)) %>% # have a CC
  filter(!is.na(strain)) %>% # assigned strain
  select(-Run, -`Sample Name`, -Sample_name, -Bytes, -create_date, -Bases,
         -version, -ReleaseDate, -AvgSpotLen, -`DATASTORE filetype`) %>%
  distinct()
```

singlecopiness
```{r singlecopiness}
all86k$issinglecopy <-  with(all86k, ifelse(PCN < 2, T, F))
```

```{r set2.5 making, for use & backup}
set2.5 <- all86k %>%
  filter(!is.na(mecA)) %>% # is def SA & we have sccmec
  filter(!is.na(PCN)) %>% # gold, SA, and has pt181
  filter(!is.na(year)) %>% # valid date, is SA
  filter(!is.na(CC)) %>% # have a CC
  filter(!is.na(strain)) %>%
  select(-Run, -`Sample Name`, -Sample_name, -Bytes, -create_date, -Bases,
         -version, -ReleaseDate, -AvgSpotLen, -`DATASTORE filetype`) %>%
  distinct()
```

## calculating depth for non-gold integrated isolates (n = 9)
locating the samples
```{r locate in df}
avgdepths %>%
  filter(Experiment == "SRX7671590" |
           Experiment == "SRX10154032" | 
           Experiment == "ERX458901" |
           Experiment == "ERX3560180" |
           Experiment == "SRX886929" |
           Experiment == "SRX4814892" |
           Experiment == "ERX2285900" |
           Experiment == "ERX2285874" |
           Experiment == "ERX1834600") %>%
  distinct(Experiment) # they are all in "avgdepths" (no stragglers)
```

get the depths for the chrom, pT181, & NA contigs for non-gold integrated isolates
```{r df with depths & positions}
ngId <- left_join(avgdepths, geneID,
                        by = c("Experiment", "contig")) %>%
  filter(Experiment == "SRX7671590" |
           Experiment == "SRX10154032" | 
           Experiment == "ERX458901" |
           Experiment == "ERX3560180" |
           Experiment == "SRX886929" |
           Experiment == "SRX4814892" |
           Experiment == "ERX2285900" |
           Experiment == "ERX2285874" |
           Experiment == "ERX1834600") %>%
  select(-gene) %>%
  distinct()
```

```{r calculate PCN}
# get avg for each location for each sample
ratioNGi <- ngId %>% 
  group_by(Experiment, loc) %>% 
  summarise(meandepth = mean(avgDepth))

untidyNGi <- pivot_wider(ratioNGi, names_from = loc,
                      values_from = meandepth)

# adding ratio of plasmid to chromosome
untidyNGi$PCN <- (untidyNGi$pT181 / untidyNGi$chrom)

untidyNGi
```
PCN is solid estimator, even for non-gold isolates!!!
range 1.01 - 1.39


# PCN in set2.5
```{r pcn histogram}
set2.5 %>%
  filter(PCN > 2) %>%
  ggplot(aes(x = PCN)) +
  geom_histogram()
```


```{r pcn stats}
set2.5backup %>%
  filter(PCN > 2) %>%
  select(PCN) %>%
  summary()

x1 <- set2.5backup %>%
  filter(PCN > 2) %>%
  select(PCN)

quantile(x1$PCN,
         probs=c(0.025,0.975))

IQR(x1$PCN)
```

```{r recent MC pT181}
set2.5 %>%
  filter(as.numeric(year) >= 1995) %>%
  filter(PCN > 2) %>%
  mutate(top9strain = fct_lump_n(strain,9)) %>%
  filter(top9strain != "Other") %>%
  ggplot(aes(x = as.numeric(year), y = PCN)) +
  geom_point() +
  facet_wrap(~top9strain, ncol = 3) +
  labs(title = "multicopy PCN by strain (top 9)") +
  theme_bw()
```

# Singlecopiness by strain
```{r chi square - sc x strain}
set2.5Schi <- set2.5 %>%
  group_by(strain, issinglecopy) %>%
  mutate(total = n()) %>%
  select(strain, issinglecopy, total) %>%
  distinct() %>%
  pivot_wider(id_cols = strain,
              names_from = issinglecopy,
              values_from = total) %>%
  mutate_all(~replace(.,is.na(.),0))

Ms <- as.table(rbind(set2.5Schi$`TRUE`, set2.5Schi$`FALSE`))
dimnames(Ms) <- list(copy = c("single", "multi"),
                    strain = c(set2.5Schi$strain))

XsqS <- chisq.test(Ms)

XsqS$residuals  # Pearson residuals
XsqS$stdres
XsqS$p.value
XsqS$statistic
```

```{r prevalence SC vs MC in pre & post}
set2.5 %>%
  group_by(prepost, issinglecopy) %>%
  mutate(scTime = n()) %>%
  select(prepost, issinglecopy, scTime) %>%
  distinct()

# postSC %
1848/(1848+428)

# preSC %
15 / (15+169)
```


## fig. 6B
```{r figure: sc x strain (Fig. 6B)}
set2.5Schi %>%
  filter(`TRUE`+`FALSE` > 1) %>%
  mutate(ratio = `TRUE`/`FALSE`) %>%
  left_join(., set2.5backup, by = "strain") %>%
  ggplot(aes(x = forcats::fct_reorder(strain, ratio, .fun = median),
         fill = issinglecopy)) +
  geom_bar(position = "fill")  +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Strain",
       y = "Proportion of Samples",
       fill = "PCN") +
  scale_fill_manual(values = c25,
                    labels = c("multicopy", "single-copy"))
```

# where is the 1990s sampling gap?
```{r}
set2.5 %>%
  filter(as.numeric(year) < 1998 & as.numeric(year) > 1990) %>%
  select(sample, type, strain, year)

all86k %>%
  filter(as.numeric(year) < 1997 & as.numeric(year) > 1993) %>%
  select(sample, type, strain, year)
```
No samples in set2.5k collected 1992-1996
Also there is a composition difference before/after this gap
staph is continuously collected still


# PCN for type III SCCmec
```{r type iii stats}
set2.5backup %>%
  filter(type == "III") %>%
  select(PCN) %>%
  summary()
```

```{r strain27 info}
set2.5backup %>%
  filter(issinglecopy == T) %>% # 1,863 rows
  filter(strain == "S99.5_27") # 1,131 rows
```

```{r what sccmec types s27}
set2.5backup %>%
  filter(strain == "S99.5_27") %>%
  select(mecA, type) %>%
  distinct()
```


# Integration position

## integration in complete genomes

```{r}
complot <- left_join(completeover, completeblast, by = "sseqid") %>%
  ggplot(aes(xmin = sstart, xmax = send, y = sseqid)) +
  geom_gene_arrow()

complot

#ggexport(complot, filename = "complot.pdf")
```

```{r}
comsite <- left_join(completeover, completeblast, by = "sseqid") %>%
  mutate(pos = ifelse(send < 4440, "maybefree", "int")) 

comsite %>%
  ggplot(aes(xmin = sstart, xmax = send, y = sseqid,
             color = pos
             )) +
  geom_gene_arrow()

# Complete genomes from 337 set have GCA names
dnaA_start_coords$samplematch <- dnaA_start_coords$contig %>%
  str_remove("GCA_.*_")

dnaA_start_coords %>%
  select(genome) %>%
  distinct()

dnaA_start_coords %>%
  select(contig) %>%
  distinct()

comsite$samplematch <- comsite$sseqid %>%
  str_remove("N._")

left_join(dnaA_start_coords, comsite, by = "samplematch") %>%
  filter(!is.na(sseqid)) %>%
  select(samplematch) %>%
  distinct() # 14 dnaA-1st complete genomes have pT181
```
Complete genomes from 337 set have GCA names
There are 14 of the dnaA start genomes that are in the set with pT181


```{r}
inner_join(dnaA_start_coords, comsite, by = "samplematch") %>%
  select(samplematch, length.y:pos) %>%
  distinct() %>%
  ggplot(aes(xmin = sstart, xmax = send, y = samplematch,
             color = pos
             )) +
  geom_gene_arrow()
```

```{r}
position14 <- inner_join(dnaA_start_coords, comsite, by = "samplematch")
```

```{r}
geneset <- position14 %>%
  filter(sseqid == "NZ_CP015447.2" |
           sseqid == "NZ_CP029172.1" |
           sseqid == "NZ_CP029685.1" |
           sseqid == "NZ_CP022902.1" |
           sseqid == "NZ_CP058312.1" |
           sseqid == "NZ_LS483309.1" |
           sseqid == "NZ_LR027874.1") %>%
  filter(start > 44000 & start < 100000) 

#geneset$mutName <- ifelse(str_detect(geneset$symbol, "MecA") == TRUE, 'mecA',
 #                           ifelse(str_detect(geneset$symbol,
  #                                            "tetracycline efflux MFS transporter Tet") == TRUE, "tetK",
   #      ifelse(str_detect(geneset$symbol, "transposase") == TRUE, "transposase",
    #            ifelse(str_detect(geneset$symbol, "replication initiation protein") == TRUE,"repC", 
     #                  ifelse(geneset$symbol, "plasmid recombination protein") == TRUE, "pre", 
      #                        ifelse(geneset$symbol,"recombinase") == TRUE,"recombinase", 
       #                "other"))))
geneset %>%
  ggplot(aes(xmin = start, xmax = end, y = samplematch,
             fill = symbol,
             color = symbol
             )) +
  geom_gene_arrow()
```


# CCN: verifying that core contigs are centered at 1 & origin isn't

```{r calculate chrom avg}
# get avg for each location for each sample
chromdepths <- alldepths1 %>%
  filter(loc == "chrom") %>%
  group_by(Experiment) %>% 
  mutate(ChromAvg = mean(avgDepth))

chromdepths$CCN <- chromdepths$avgDepth / chromdepths$ChromAvg

chromdepths %>%
  filter(CCN < 2) %>%
  ggplot(aes(x = CCN)) +
  geom_histogram()
```

So it looks like a pretty even distribution!

need to go back to the genes to approximate oriC location
Defined by co-localization of 4 genes (Lancet paper, 2001)
- rpmH = g03298
- dnaA = g00489
- dnaN = g00392
- recF = g00595

```{r ID near origin}
fo1 <- left_join(chromdepths, geneID) %>%
  select(Experiment:avgDepth, loc:gene)

fo2 <- left_join(chromdepths, stragger_lawfulpos) %>%
  select(Experiment:avgDepth, loc:gene)

findOri <- full_join(fo1, fo2)

findOri$nearOri <- with(findOri, ifelse(gene == "g03298" |
           gene == "g00489" |
           gene == "g00392" |
           gene == "g00595",
           T, F))
```

I think the NAs have been generated by the join.

```{r Fig. 11 chrom plot near origin vs not}
findOri %>%
  select(-gene) %>%
  filter(!is.na(nearOri)) %>% #only stuff that had an identifiable gene
  distinct() %>% # de-replicate contigs
  filter(CCN < 3) %>% # for readability
  ggplot(aes(x = CCN, fill = nearOri)) +
  geom_histogram() +
  theme_bw()
```


One sample t-test (data is normally distributed)
```{r origin ttest & plot}
contigNearOri <- findOri %>%
  filter(nearOri ==T)

t.test(contigNearOri$CCN, mu = 1)

ggplot(data = contigNearOri,
       aes(x = CCN)) +
  geom_histogram()

contigNearOri$CCN %>%
  mean()

contigNearOri$CCN %>%
  sd()
```
For contigs containing/right next to origin, CCN = 1.15
This is significant, True mean not 1 (like it is/should be for chrom overall)
Distribution isn't actually as tight as the type III distribution is

# looking at early isolates
```{r early strain breakdown}
# assess sampling bias
all86k %>%
  filter(prepost == "pre") %>% # looking at early samples
  filter(!is.na(strain)) %>%
  filter(!is.na(CC)) %>%
  filter(!is.na(mecA)) %>%
  filter(rank == "gold") %>%
  select(sample, strain, prepost, year) %>%
  distinct() %>% # 242 pre-1995 isolates
  #filter(strain == "S99.5_2") # 225 strain 2
  filter(strain != "S99.5_2") #17 not strain 2

# with pT181
all86k %>%
  filter(prepost == "pre") %>% # looking at early samples
  filter(!is.na(pT181contigSA)) %>%
  filter(!is.na(strain)) %>%
  filter(!is.na(CC)) %>%
  filter(!is.na(mecA)) %>%
  filter(rank == "gold") %>%
  select(sample, strain, prepost, year) %>%
  distinct() %>% # 242 pre-1995 isolates
  filter(strain == "S99.5_2") # 182 strain 2
  # filter(strain != "S99.5_2") #2 not strain 2
```
this is for HIGHTEST QUALITY SA samples

```{r assessing early samples, all quality}
# all isolates
all86k %>%
  filter(prepost == "pre") %>%
  select(sample, year, strain, pT181contigSA) %>%
  distinct() %>% # 494 isolates collected before 1995
  group_by(strain) %>%
  mutate(preIsolatesInStrain = n()) %>%
  select(strain, preIsolatesInStrain) %>%
  distinct() # 249 pre-1995 isolates (all quality, all pT181 p/a) are in strain 2

# pT181-containing
all86k %>%
  filter(prepost == "pre") %>%
  filter(!is.na(pT181contigSA)) %>%
  select(sample, year, strain, pT181contigSA) %>%
  distinct() %>% #241 early isolates have pT181
  group_by(strain) %>%
  mutate(preIsolatesInStrain = n()) %>%
  select(strain, preIsolatesInStrain) %>%
  distinct() # 198 pre-1995 pT181-containing isolates are in strain 2

# non-s99.5_2 pT181 appears in the 80s and 90s
all86k %>%
  filter(prepost == "pre") %>%
  filter(!is.na(pT181contigSA)) %>%
  select(sample, year, strain, pT181contigSA) %>%
  filter(strain != "S99.5_2")
```

# year of collection figure: Fig. 1
Fig. 1, x = year of collection, y = sample count
A: pT181 P/A
B: pT181 isolates by strain

```{r Fig. 1}
all86k %>%
  filter(rank == "gold") %>%
  filter(!is.na(year)) %>%
  filter(!is.na(type)) %>%
  filter(!is.na(CC)) %>%
  filter(!is.na(strain)) %>%
  select(sample, year, pT181contigSA) %>%
  distinct() %>% 
  mutate(PT181presenceAbsence = ifelse(is.na(pT181contigSA), "no p1T81", "pT181+")) %>%
  ggplot(aes(x = as.numeric(year), fill = PT181presenceAbsence)) +
  geom_histogram() +
  theme_bw() +
  labs(y = "Sample count",
       x = "Year of Collection",
       fill = "pT181 presence/absence"
       ) +
  theme(axis.text = element_text(size = 13),
        axis.title = element_text(size = 13))

set2.5 %>%
  mutate(top7strain = fct_lump_n(strain,7)) %>%
  ggplot(aes(x = as.numeric(year), fill = top7strain)) +
  geom_histogram() +
  labs(y = "Sample count",
       x = "Year of Collection",
       fill = "Strain") +
  theme_bw() +
  scale_fill_manual(values = top7strain2.5) +
  theme(axis.text = element_text(size = 13),
        axis.title = element_text(size = 13))
```

```{r how many strains was pT181 detected in?}
all86k %>%
  select(sample, pT181contigSA, strain) %>%
  filter(!is.na(pT181contigSA)) %>%
  distinct() %>% # confirmed 7,927 rows
  select(strain) %>%
  distinct() #45 rows, 1 of them is an NA -> 44 strains
```


regression for prevalence
```{r logit regr prevalence}
# get data ready for regression on plasmid prevalence
lmprev <- all86k%>%
  filter(!is.na(year)) %>%
  filter(rank == "gold") %>%
  filter(!is.na(strain)) %>%
  filter(!is.na(type)) %>%
  filter(!is.na(CC)) %>%
  select(sample, pT181contigSA, year) %>%
  distinct() %>%
  select(pT181contigSA, year) %>%
  group_by(pT181contigSA, year) %>%
  mutate(haspt181 = ifelse(is.na(pT181contigSA), 0, 1)) 

# regression
logit.prev <- glm(haspt181 ~ as.numeric(year), data = lmprev,
    family = binomial(link = "logit")) 

summary(logit.prev)
```

# Fig. 6A: sc x time figure
```{r prop sc by time (Fig. 6A)}
scatterSC <- set2.5backup %>%
  select(sample, issinglecopy, year) %>%
  mutate(copynumber = ifelse(issinglecopy==T, "Single-copy", "Multicopy")) %>%
  group_by(copynumber, year) %>%
  mutate(countCNyear = n()) %>%
  pivot_wider(names_from = copynumber, values_from = countCNyear) %>%
  select(-sample, -issinglecopy) %>%
  distinct() %>%
  summarise_each(funs(sum(., na.rm = TRUE))) %>%
  mutate(propSC = `Single-copy`/(`Single-copy` + Multicopy)) %>%
  ggplot(aes(x = as.numeric(year), y = propSC)) +
  geom_point() +
  ylim(0,1) +
  labs(x = "Year of Collection",
       y = "Proportion Single-copy") +
  theme_bw() +
  geom_smooth(method = "lm", se = F)

scatterSC
```

```{r Fig. 6A reg}
dataforproplm <- set2.5backup %>%
  select(sample, issinglecopy, year) %>%
  mutate(copynumber = ifelse(issinglecopy==T, "Single-copy", "Multicopy")) %>%
  group_by(copynumber, year) %>%
  mutate(countCNyear = n()) %>%
  pivot_wider(names_from = copynumber, values_from = countCNyear) %>%
  select(-sample, -issinglecopy) %>%
  distinct() %>%
  summarise_each(funs(sum(., na.rm = TRUE))) %>%
  mutate(propSC = `Single-copy`/(`Single-copy` + Multicopy))

lm(propSC ~ as.numeric(year), data = dataforproplm) %>%
  summary()
```

```{r added 2025-09-15: logistic regression on SC}
set2.5backup %>%
  select(sample, issinglecopy, year)

logit.sc <- glm(issinglecopy ~ as.numeric(year), data = set2.5backup,
    family = binomial(link = "logit")) 

summary(logit.sc)
```

Check - how many isolates were collected before 1950?
```{r pre-1950}
all86k %>%
  filter(year < 1950) %>%
  select(sample, strain, year)
```
8 isolates before 1950
strain breakdown: 3 NA, 3 S21, 1 S5, 1 S15


# Thresholding: why >= 4400bp for present

```{r Fig. 9 understanding blast hit distribution}
summary(totallength)

totallength %>%
  ggplot(mapping = aes(x = length)) + 
  geom_histogram(binwidth = 100) +
  xlab("sum of blast hit length") +
  theme_bw()

totallength %>%
  filter(length >= 3500 & length <= 4800) %>%
  ggplot(mapping = aes(x = length)) + 
  geom_histogram(binwidth = 100) + 
  scale_x_continuous(
    breaks = c(3500,3600,3700,3800,3900,4000,4100,4200,
               4300,4400,4500,4600,4700,4800),
    labels = c(3500,3600,3700,3800,3900,4000,4100,4200,
               4300,4400,4500,4600,4700,4800))

totallength %>%
  filter(length >= 4200 & length <= 4600) %>%
  ggplot(mapping = aes(x = length)) + 
  geom_histogram(bins = 30) +
  xlab("sum of blast hit length") +
  theme_bw()

totallength %>%
  filter(length >= 4400 & length <= 4500) %>%
  ggplot(mapping = aes(x = length)) + 
  geom_histogram(bins = 25) 
```

# Coefficient of variation in read depth

```{r load the coevar data}
cvD <- read_delim("10golddepths.forcoevar.tsv", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)

cvD <- cvD %>%
  rename("sample" = "X1",
         "contig" = "X2",
         "position" = "X3",
         "depth" = "X4")
```

```{r get cvD statistics}
cvDstat <- cvD %>% 
  group_by(sample, contig) %>% 
  summarise(avg = mean(depth),
            var = var(depth),
            length = length(depth),
            coevar = sd(depth)/mean(depth))
```

```{r}
cvDstat %>%
  ggplot(aes(x = coevar)) +
  geom_density() +
  facet_wrap(~sample)
```

```{r}
cvDstat %>%
  ggplot(aes(x = sample, y = avg)) +
  geom_point() +
  geom_boxplot() +
  scale_y_sqrt()

cvD %>%
  filter(sample == "DRX203040") %>%
  ggplot(aes(x = position, y = depth)) +
  geom_line() +
  facet_wrap(~contig, ncol = 1)
```

## join with gene positions to be able to pick apart core / pT181 / other contigs
```{r load contig lengths}
depth_pcn_allSA <- read_table2("20240321-depth-pcn-allSA.txt", 
    col_names = FALSE)

depth_pcn_allSA <- depth_pcn_allSA %>%
  rename("Experiment" = "X1", "contigOGname" = "X2",
         "avgDepth" = "X3", "contigLen" = "X4")
```

```{r}
alldepthsC <- merge(alldepths1, depth_pcn_allSA, by = c("Experiment", "avgDepth"))

left_join(cvDstat, alldepthsC, by = c("sample" = "Experiment",
                                     "contig" = "contigOGname"))

left_join(cvDstat, alldepthsC, by = c("sample" = "Experiment",
                                     "contig" = "contigOGname")) %>%
  filter(!is.na(loc)) %>%
  ggplot(aes(x = sample, y = avgDepth, color = loc)) +
  geom_point() +
  scale_y_sqrt()
```

```{r coevar by length of contig}
left_join(cvDstat, alldepthsC, by = c("sample" = "Experiment",
                                     "contig" = "contigOGname")) %>%
  mutate(lenCat = ifelse(length < 5000, "short",
                         ifelse(length < 40000, "med", "long"))) %>%
  ggplot(aes(x = coevar)) +
  geom_density() +
  facet_wrap(~lenCat, nrow = 1)
```
length of contig does NOT impact coefficient of variation


```{r Fig. 12 coevar by loc}
left_join(cvDstat, alldepthsC, by = c("sample" = "Experiment",
                                     "contig" = "contigOGname")) %>%
  ggplot(aes(x = coevar)) +
  geom_density() +
  facet_wrap(~loc, nrow = 1) +
  theme_bw()
```
loc doesn't matter for coevar, curves look pretty much the same


```{r coevar by loc significance}
forW <- left_join(cvDstat, alldepthsC, by = c("sample" = "Experiment",
                                     "contig" = "contigOGname")) %>%
  select(sample, coevar, loc) %>%
  filter(!is.na(loc))

wilcox.test(coevar ~ loc, data = forW)
# location not predictive of coefficient of variation by wilcox rank sum
```

Integration position results show
- integration both up and downstream of mecA
- different rotation positions
- integration forward / reverse

# GLMM code (strain/time/instrument/center) is on server

# Linearization position - what effect does it have on snippy results?
we need to check and see if what I'm doing for linearization is impacting estimates of mutation rate and 

load the start positions
```{r saccver & rename}
# most samples
startMost <- read_delim("technicalLin/saccversstart.txt", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)

startMost <- startMost %>%
  rename("sample.contig" = "X1",
         "start" = "X2",
         "start-1" = "X3")

# nonSA samples
startnonSA <- read_delim("technicalLin/nonSA-saccversstart.txt", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)

startnonSA <- startnonSA %>%
  rename("sample.contig" = "X1",
         "start" = "X2",
         "start-1" = "X3") %>%
  mutate(isSA = F) %>%
  filter(sample.contig !="AY672108.1" & # filtering out the cloning vectors
          sample.contig !="JX893516.1")

# stragglers
startstrag <- read_delim("technicalLin/strag-saccversstart.txt", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)
startstrag <- startstrag %>%
  rename("sample.contig" = "X1",
         "start" = "X2",
         "start-1" = "X3")

startSA <- full_join(startMost, startstrag) %>%
  mutate(isSA = T)


startSA$experiment <- startSA$sample.contig %>%
  str_remove("-cont.*.$")
# this is for use later

startall <- full_join(startSA, startnonSA) %>%
  select(-experiment)
```

```{r join with snippy for lindf}
lindf <- left_join(startall, df2, by = c("sample.contig" = "SAMPLE", "isSA")) %>%
  select(-"start-1") # remove unneeded column
```

```{r understanding lindf}
summary(lindf$start)

unique(lindf$start)

lindf %>%
  mutate(startcat = ifelse(start == 1,
                           "one", ifelse(start > 4439, "longish",
                                         "inpT181"))) %>%
  select(Experiment, startcat) %>%
  distinct() %>%
  select(startcat) %>%
  table()

lindf %>%
  mutate(startcat = ifelse(start == 1,
                           "one", ifelse(start > 4439, "longish",
                                         "inpT181"))) %>%
  #ggplot(aes(x = as.numeric(POS), fill = startcat)) +
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram() +
  facet_wrap(~startcat, scales = "free", ncol = 1)
```

just checking sample counts
```{r}
lindf %>%
  distinct(sample.contig)
```
7,988 rows -- this has got everything

what has 0 snps relative to reference?
```{r}
lindf%>%
  filter(is.na(TYPE))
```
3 isolates + the DRX100326 reference
INTERESTING that the 3 that are the same are nonSA that have long starts

DRX100326 starts at 3084 relative to blast query
what else does?

```{r just DRX100326 linearization position isolate snps}
lindf %>%
  filter(start == 3084) %>%
  # distinct(sample.contig) #1,366 have same linearization position
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```
ummmmmm oh no
if we only look at stuff with the same linearization position as the snippy reference, there are no insertions, only a handfull of deletions, and a few complex mutations
its mostly snps
BUT it is still the spikes that we have previously seen

```{r}
lindf %>%
  filter(start == 3084) %>%
  select(POS, TYPE, ALT) %>%
  distinct() # 181 rows
```

```{r PCN check}
left_join(lindf, set2.5backup, by = c("Experiment" = "sample")) %>%
  filter(!is.na(PCN)) %>%
  filter(start == 3084) %>%
  select(Experiment, PCN, issinglecopy) %>%
  distinct() %>%
  #summary(PCN)
  ggplot(aes(x = PCN)) +
  geom_histogram() 
```
This linearization position is present for the full range of PCN

```{r}
left_join(lindf, set2.5backup, by = c("Experiment" = "sample")) %>%
  filter(!is.na(PCN)) %>%
  filter(start == 3084) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram() +
  facet_wrap(~issinglecopy, ncol = 1)
```
Both single and multicopy pT181 with same lin position have complex & snps before pre

## checking on pre-tetK mutation spike
I want to check on thre pre-tetK mutants
```{r start pos for tet spike samples}
lindf %>%
  filter(POS > 810 & POS < 820) %>%
  # there is an outlier that was linearized outside of the canonical plasmid length - exclude for visualization
  filter(start < 4500) %>%
  ggplot(aes(x = as.numeric(start))) +
  geom_histogram()
```
The vast majority were linearized ~3635
Where is this in the plasmid?
Can we tell this for all of them? Actually maybe not because this is sstart not qstart
but for qstart its 1
find what the annotation positions would be for the query sequence

"NC_001393.1 Staphylococcus aureus plasmid pT181, complete sequence"
This may likely be from a complete genome, not an SRA assembly

if we get the annotations from something else with start 1, then get that annotation
we'd be good
```{r}
lindf %>%
  filter(start == 1) %>%
  distinct(sample.contig)
```

## trends for early starters
let's define this as plasmids linearized before DRX100326
```{r pre-DRX ref starters}
lindf %>%
  filter(start < 3084) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```

## correlation between start & pos
```{r Fig. S3}
# DRX100326 starts at 3084 relative to blast query
lindf <- lindf %>%
  mutate(aheadby = start - 3084)

lindf %>%
  filter(start < 4439) %>%
  filter(!is.na(TYPE)) %>%
  ggplot(aes(x = start, y = POS, color = TYPE)) +
  geom_point() +
  labs(x = "hitStart",
       y = "Mutation Position") +
  theme_bw()

lindf %>%
  filter(start < 5000) %>%
  ggplot(aes(x = aheadby, y = POS, color = TYPE)) +
  geom_point()
```
The mutations that occur in the spike positions occur regardless of where the start of the contig is --> the spikes are not due to a particular start position or rotation bringing in new genetic material into what we are calling pT181

```{r}
lindf %>%
  #filter(start < 10000) %>%
  #distinct(sample.contig) # 7,941
  filter(start < 5000) %>%
  distinct(sample.contig, aheadby) %>% # 7,919
  ggplot(aes(x = aheadby)) +
  geom_histogram()
```
Most start at the same place or soon after DRXref -- this looks like late rep. Some are in earlier rep
This is a little bit of a result, by maybe belongs more in methods


```{r}
lindf %>%
  filter(start < 5000) %>%
  filter(TYPE != "snp") %>%
  ggplot(aes(x = aheadby, y = POS, color = TYPE)) +
  geom_point()
```


```{r linear model for position}
lm(POS ~ start, data = lindf) %>%
  summary()
```


## learning about samples with indels
```{r}
indelSamples <- lindf %>%
  filter(TYPE == "ins" | TYPE == "del") %>%
  select(sample.contig) %>%
  distinct()

left_join(indelSamples, lindf) %>%
  # some contigs are long, exclude those
  filter(start < 4500) %>%
  select(sample.contig, start) %>%
  distinct()%>%
  ggplot(aes(x = start)) +
  geom_histogram()
```
Most are linearized at ~3650
but there is some variation

```{r reflin vs other - indel samples}
left_join(indelSamples, lindf) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```

YOU CAN'T COMPARE SSTART TO DRXREF START!
add a column to dataframe that would let you do that

DRX100326 starts at 3084 relative to blast query

```{r}
lindf %>%
  select(sample.contig, aheadby) %>%
  distinct() %>%
  filter(aheadby < 4000) %>%
  ggplot(aes(x = aheadby)) +
  geom_histogram()
```
most pT181 are linearized a little later in the plasmid than the snippy ref
This is in tetK and explains why I was having problems doing in silico PCR for it
DRX snippy ref is linearized right before tetK

Bactopia annotation for DRX ref
125	979	.	+	0	tet(K) --- TETK
1165	2406	.	+	0	plasmid recombination protein --- PRE
2592	2693	.	+	0	product=hypothetical protein
2933	3877	.	+	0	replication initiation protein --- REPC

Bakta annotation for norm 
Prodigal	CDS	45	794	.	+	0	Name=Replication initiation protein
Bakta	CDS	862	912	.	+	0	Name=OrfE
Prodigal	CDS	956	2335	.	+	0	Name=tetracycline efflux MFS transporter Tet(K)
Prodigal	CDS	2521	3762	.	+	0	ID=JAIFDI_00020;Name=Plasmid recombination enzyme type 3 --- pre

BLAST+	oriC	3763	4288	.	?	.	Name=origin of replication

Prodigal	CDS	3948	4049	.	+	0	Name=CPBP family intramembrane metalloprotease

Infernal	ncRNA	4100	4194	1.3e-14	-	.gene=ctRNA_pT181;product=ctRNA


### look at long contig mutation patterns

```{r}
gt4439 <- alldepthsC %>%
  filter(loc == "pT181") %>%
  filter(contigLen > 4439)

left_join(gt4439, lindf, by = "Experiment") %>%
  #distinct(Experiment) # 4,615 rows
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()

left_join(gt4439, lindf, by = "Experiment") %>%
  #filter(start < 2*10^4) %>%
  #filter(start < 5000) %>%
  filter(start > 3500 & start < 4000) %>%
  ggplot(aes(x = start)) +
  geom_histogram()
```
Most have start ~3635 area
DRXref linearizes at 3084 in its own position (this is before tetK)

snippy uses *normalized DRXref*
positions are relative to normalized pT181
gene plot is correct - those are the correct locations


## look for mutations exclusive to long contigs
split dataset up into <= 4439bp and >= 4440bp
then, look for mutations present in 4440+ dataset that aren't in short one

```{r}
pT181len <- alldepthsC %>%
  filter(loc == "pT181") %>%
  select(Experiment, contig, contigLen)
```

```{r splitting df short/long}
longMut <- left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short")) %>%
  filter(lengthCat == "long") %>%
  group_by(POS, TYPE) %>%
  mutate(countOfPosT = n()) %>%
  filter(countOfPosT > 1) %>%
  select(POS, TYPE, REF, ALT) %>%
  distinct()

shortMut <- left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short")) %>%
  filter(lengthCat == "short") %>%
  select(POS, TYPE) %>%
  mutate(countOfPosT = n()) %>%
  filter(countOfPosT > 1) %>%
  distinct()
```

I'm curious about the short contigs -- do they have any indels?
```{r short mut}
shortMut %>%
  ggplot(aes(as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```
YES, they do -- scattered across the plasmid
The presence of indels should not immediately trigger red flags

```{r all mut on short contigs}
left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short")) %>%
  filter(lengthCat == "short") %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```
Spikes are in the same place
don't see a before tetK spike in quite the same place but there is a deletion there
Can't really have long mut for the short contigs

```{r long mut}
longMut %>%
  ggplot(aes(as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```



```{r only 2+ long}
anti_join(longMut, shortMut, by = c("POS", "TYPE")) %>%
  #View()
  filter(TYPE != "snp") %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram()
```


where are the spikes?
```{r}
aj.df <- anti_join(longMut, shortMut, by = c("POS", "TYPE")) %>%
  filter(TYPE != "snp")

left_join(aj.df, lindf, by = c("POS", "TYPE", "REF", "ALT")) %>%
  ungroup() %>%
  distinct(sample.contig) # 1,309 rows


left_join(aj.df, lindf, by = c("POS", "TYPE", "REF", "ALT")) %>%
  ungroup() %>%
  group_by(sample.contig) %>%
  mutate(occurance = n()) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram() +
  facet_wrap(~occurance, ncol = 1, scales = "free")
```
All of the isolates with 3 spikes (n = 5) have the spikes  at 814, 3197, & 3672
2 spikes (n = 19)
1 spike (n = 1,285)


```{r}
left_join(aj.df, lindf, by = c("POS", "TYPE", "REF", "ALT")) %>%
  ungroup() %>%
  group_by(sample.contig) %>%
  mutate(occurance = n()) %>%
  left_join(., set2.5backup, by = c("Experiment" = "sample")) %>%
  filter(!is.na(PCN)) %>%
  select(POS:ALT, Experiment, PCN, issinglecopy) %>%
  ungroup() %>%
  #distinct(Experiment) # 593 rows
  select(Experiment, PCN) %>%
  distinct() %>%
  View() # some are singlecopy, some are multicopy
```


```{r}
left_join(aj.df, lindf, by = c("POS", "TYPE", "REF", "ALT")) %>%
  ungroup() %>%
  group_by(sample.contig) %>%
  mutate(occurance = n()) %>%
  ungroup() %>%
  left_join(., set2.5backup, by = c("Experiment" = "sample")) %>%
  filter(!is.na(PCN)) %>%
  select(POS:ALT, Experiment, PCN, issinglecopy) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram() +
  facet_wrap(~issinglecopy, ncol = 1, scales = "free")
```

```{r}
left_join(aj.df, lindf, by = c("POS", "TYPE", "REF", "ALT")) %>%
  ungroup() %>%
  group_by(sample.contig) %>%
  mutate(occurance = n()) %>%
  ungroup() %>%
  left_join(., set2.5backup, by = c("Experiment" = "sample")) %>%
  select(POS:ALT, Experiment, PCN, issinglecopy) %>%
  filter(PCN > 50) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram() 
```


Let me write about this

if linearization is causing extra sequence to get pulled in, we should see a higher rate of indels and complex mutations in long contigs compared to short contigs

we can check that statistically

```{r getting cat total}
left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short")) %>%
  ungroup() %>%
  select(Experiment, lengthCat) %>%
  distinct() %>%
  select(lengthCat) %>%
  table()
```

```{r prop mut in cat}
left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short"),
         catTotal = ifelse(lengthCat == "long", 4609, 1114)) %>%
  ungroup() %>%
  select(Experiment, lengthCat, catTotal, POS, TYPE, ALT) %>%
  group_by(lengthCat, POS, TYPE, ALT) %>%
  mutate(countInCat = n(),
         propCat = countInCat/catTotal)
```

```{r}
left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short"),
         catTotal = ifelse(lengthCat == "long", 4609, 1114)) %>%
  ungroup() %>%
  select(Experiment, lengthCat, catTotal, POS, TYPE, ALT) %>%
  group_by(lengthCat, POS, TYPE, ALT) %>%
  mutate(countInCat = n(),
         propCat = countInCat/catTotal) %>%
  ungroup() %>%
  select(lengthCat:countInCat) %>%
  distinct() # 584 rows

chiPrep <- left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short"),
         catTotal = ifelse(lengthCat == "long", 4609, 1114)) %>%
  ungroup() %>%
  select(Experiment, lengthCat, catTotal, POS, TYPE, ALT) %>%
  group_by(lengthCat, POS, TYPE, ALT) %>%
  mutate(countInCat = n(),
         propCat = countInCat/catTotal) %>%
  ungroup() %>%
  select(lengthCat, POS:propCat) %>%
  distinct()

chiPrep$allele <- paste0(chiPrep$POS,"-", chiPrep$TYPE,"-", chiPrep$ALT)
```

```{r}
chiPrep %>%
  ggplot(aes(x = allele, y = countInCat)) +
  geom_col()
```



```{r chi further set up}
chiT <- chiPrep %>%
  select(allele, lengthCat, countInCat) %>%
  pivot_wider(names_from = lengthCat, values_from = countInCat) %>%
  mutate_all(~replace(.,is.na(.),0)) 

chi.table <- as.table(rbind(chiT$long, chiT$short))

dimnames(chi.table) <- list(
  length = c("long", "short"),
  allele = c(chiT$allele)
  )

chi.table

XsqT <- chisq.test(chi.table)

XsqT$residuals  # Pearson residuals
XsqT$stdres
XsqT$expected
XsqT$observed
```


```{r trying again}
distinctA <- left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  mutate(lengthCat = ifelse(as.numeric(contigLen) > 4439, "long", "short"),
         catTotal = ifelse(lengthCat == "long", 4609, 1114)) %>%
  ungroup() %>%
  select(POS:Experiment, lengthCat, catTotal) %>%
  group_by(POS, TYPE, lengthCat) %>%
  mutate(countPOStypeCat = n()) %>%
  filter(TYPE != "snp") %>%
  ungroup() %>%
  mutate(catWithout = catTotal - countPOStypeCat) %>%
  select(POS, TYPE, lengthCat:catWithout) %>%
  distinct()

distinctA$pos.type <- paste0(distinctA$POS,"-", distinctA$TYPE)
```


```{r}
chiset <- distinctA %>%
  select(-POS, -TYPE, -catTotal, -catWithout) %>%
  pivot_wider(names_from = lengthCat,
              values_from = countPOStypeCat) %>%
  mutate_all(~replace(.,is.na(.),0)) %>%
  mutate(longWithout = 4609 - long,
         shortWithout = 1114 - short)
chiset
```


```{r chisq function}
chisq_last4cols <- function(row) {
  # Extract last 4 values of the row
  counts <- as.numeric(tail(row, 4))
  
  # Create 2x2 matrix:
  # Rows: long vs short
  # Columns: with vs without mutation
  # Order: long, longWithout, short, shortWithout
  mat <- matrix(c(counts[1], counts[3], counts[2], counts[4]), nrow = 2,
                byrow = FALSE)
  
  # Run chi-square test
  test <- chisq.test(mat)
  
  # Return p-value
  return(c(p_value = test$p.value))
}
```


```{r apply chisquare test to df}
chiset$pval <- apply(chiset, 1, chisq_last4cols)
```

```{r}
chiset %>%
  filter(pval < 0.05/113) 
```
There is nothing universal to either long or short or even at that high of a frequency

## mutations on canonical PCN isolates
```{r}
left_join(lindf, pT181len, by = "Experiment") %>%
  filter(!is.na(contigLen)) %>%
  left_join(., all86k, by = c("Experiment" = "sample")) %>%
  filter(PCN > 18 & PCN < 26) %>%
  ggplot(aes(x = as.numeric(POS), fill = TYPE)) +
  geom_histogram() +
  #facet_wrap(~start, scales = "free_y")
  facet_wrap(~contigLen, scales = "free_y")
```
Canonical PCN (18 < PCN < 26) plasmids have spikes as well
Don't really see signal for start position or contig length


# checking on All The Bacteria filtering
does it remove samples from all86k?
```{r}
all86k %>%
  select(sample, year, strain, rank, PCN, CC, pubmlst_cc.x, type) %>%
  distinct() %>%
  filter(!is.na(year)) %>% # 42,522
  filter(!is.na(strain)) %>% # 26,815
  filter(rank == "gold") %>% # 19,654
  filter(!is.na(PCN)) %>% # 2,694
  filter(!is.na(CC)) %>% # 2,490 
  #or
  #filter(!is.na(pubmlst_cc.x)) # 2,053 -- don't use this one
  filter(!is.na(type)) # 2,460
```
Yes, using AllTheBacteria removed additional genomes -- 30 samples

# position in completes

Now that there are only 14, you can manually check to see if they are on the chromosome or free or on another plasmid
```{r}
inner_join(dnaA_start_coords, comsite, by = "samplematch") %>%
  select(sseqid) %>%
  distinct(
  )

position14 <- inner_join(dnaA_start_coords, comsite, by = "samplematch")

```

```{r}
left_join(position14, X14_description_ncbi, by = "sseqid") %>%
  select(sseqid, pident:position) %>%
  distinct() %>%
  ggplot(aes(xmin = sstart, xmax = send, y = sseqid
             #color = position
             )) +
  geom_gene_arrow() +
  facet_wrap(~position, scales = "free")
```

7 integrated into chromosome
3 free
4 integrated into other plasmids

```{r}
left_join(position14, X14_description_ncbi, by = "sseqid") %>%
  select(sseqid, pident:position) %>%
  distinct() %>%
  filter(position == "chromInt") %>%
  ggplot(aes(xmin = sstart, xmax = send, y = sseqid
             )) +
  geom_gene_arrow() 
```

Need to investigate the hits that occur in CP022902 -- the far fragments look too small
confirmed (block below - real hit is 58-63k, others should not be considered pT181)
all (other) integrations occurring 55k-85k position
5 forward, 2 reverse

```{r}
left_join(position14, X14_description_ncbi, by = "sseqid") %>%
  select(sseqid, pident:position) %>%
  distinct() %>%
  filter(sseqid == "NZ_CP022902.1")
```

```{r}
left_join(position14, X14_description_ncbi, by = "sseqid") %>%
  select(sseqid, pident:position) %>%
  distinct() %>%
  filter(position == "plasInt") %>%
  ggplot(aes(xmin = sstart, xmax = send, y = sseqid
             )) +
  geom_gene_arrow() 
```

1 integrated forward, 3 integrated reverse
In 4 different plasmids (lengths 21-44kbp)

```{r}
X14_description_ncbi %>%
  filter(position == "plasInt") #%>%
  #select(description)
```

I want to get the gene positions in/around/near pT181 for the 14
this block is just for integrated
```{r}
position14 %>%
  filter(start > 55000 & end < 90000) %>%
  filter(!is.na(symbol)) %>%
  select(gene:samplematch, sseqid) %>%
  distinct() %>%
  ggplot(aes(xmin = start, xmax = end, y = sseqid,
             color = symbol,
             fill = ifelse(symbol == "tet(K)", T, F))) +
  geom_gene_arrow()
```

```{r Fig. S3: region of pT181 insertion in complete chrom}
position14 %>%
  filter(start > 55000 & end < 90000) %>%
  filter(!is.na(symbol)) %>%
  select(gene:samplematch) %>%
  distinct() %>%
  filter(symbol != "tet(K)" &
           symbol != "repC" &
           symbol != "pre") %>%
  ggplot(aes(x = symbol, fill = species_category)) + 
  geom_bar() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "gene",
       y = "# genomes found in",
       fill = "gene category")
```


# citations
```{r}
citation()
```


