---
title: "snpdists-comparison"
author: "Megan Phillips"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r loading snpdist output}
aln_dist <- read_delim("ska.noref.dist.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE) %>%
  rename("seq1" = "snp-dists 0.8.2") %>%
  column_to_rownames(var = "seq1")

map_dist <- read_delim("ska.ref.dist.tsv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE) %>%
  rename("seq1" = "snp-dists 0.8.2") %>%
  column_to_rownames(var = "seq1")
```


```{r aln matrix & long}
# Convert to matrix
aln_dist_m <- as.matrix(aln_dist)

# Extract upper triangle
aln_uppertri <- which(upper.tri(aln_dist_m), arr.ind = TRUE)

long_aln_df <- data.frame(
  sample1 = rownames(aln_dist_m)[aln_uppertri[, 1]],
  sample2 = colnames(aln_dist_m)[aln_uppertri[, 2]],
  snpdist.aln = aln_dist_m[upper.tri(aln_dist_m)]
)
# no duplicates or self-comparisons

long_aln_df
```

```{r map matrix & long}
# Convert to matrix
map_dist_m <- as.matrix(map_dist)

# Extract upper triangle
map_uppertri <- which(upper.tri(map_dist_m), arr.ind = TRUE)

long_map_df <- data.frame(
  sample1 = rownames(map_dist_m)[map_uppertri[, 1]],
  sample2 = colnames(map_dist_m)[map_uppertri[, 2]],
  snpdist.map = map_dist_m[upper.tri(map_dist_m)]
)
# no duplicates or self-comparisons

long_map_df
```


now to compare the outputs...

```{r}
compdf <- inner_join(long_map_df, long_aln_df,
           by = c("sample1", "sample2"))

compdf
```
yes okay we get 31,900,078 rows which is correct

```{r plot aln vs map snps}
compdf %>%
  ggplot(aes(x = snpdist.map, y = snpdist.aln)) +
  geom_hex() +
  geom_abline(intercept = 0, slope = 1, color = "darkorange",
              linewidth = 1) +
  geom_smooth(method = "lm")
```

# testing: are they distributed the same?
```{r wilcox}
wilcox.test(compdf$snpdist.map, compdf$snpdist.aln,
            paired = TRUE,
            alternative = "two.sided")
```
they are significantly different by wilcoxon signed rank test


```{r}
summary(compdf$snpdist.map - compdf$snpdist.aln)
```

the majority of pairs are the same by the 2 methods

```{r difference histogram}
compdf %>%
  ggplot(aes(x = snpdist.map - snpdist.aln)) +
  geom_histogram(binwidth = 1)
```
Median: same snp count
average: map < aln

RESULT: mapping to reference VERSUS reference-free alignment does not dramatically change our characterization of the plasmid

# compared to the reference genome
DRX100326

```{r get dist to ref for both methods}
DRX100326match_aln <- aln_dist_m %>%
    .[grep("DRX100326", rownames(.)), ] %>%
  as.data.frame() %>%
  rownames_to_column("isolate") %>%
  rename("snpdist_aln" = ".")

DRX100326match_map <- map_dist_m %>%
    .[grep("DRX100326", rownames(.)), ] %>%
  as.data.frame() %>%
  rownames_to_column("isolate") %>%
  rename("snpdist_map" = ".")
```

```{r}
disttoref <- full_join(DRX100326match_aln, DRX100326match_map,
                       by = "isolate") %>%
  mutate(alnlarger = snpdist_aln - snpdist_map)
disttoref
```

```{r}
disttoref %>%
  ggplot(aes(x = alnlarger)) +
  geom_histogram()
```
